\documentclass[11pt,twoside]{book}

% ── Packages ──────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathrsfs}
\usepackage{enumerate}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{geometry}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{epigraph}
% \usepackage{microtype}  % requires scalable fonts

\geometry{
  papersize={7in,10in},
  margin=1in,
  inner=1.2in,
  outer=0.8in
}

% ── Theorem environments ──────────────────────────────────────────────
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{remark}[theorem]{Remark}

\theoremstyle{remark}
\newtheorem*{notation}{Notation}
\newtheorem*{convention}{Convention}

% ── Lean 4 code listings ─────────────────────────────────────────────
\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,structure,class,instance,namespace,
    section,end,open,import,variable,where,by,have,let,show,calc,
    match,with,if,then,else,do,return,fun,forall,exists,Prop,Type,
    sorry,exact,apply,intro,rw,simp,linarith,omega,induction,cases,
    constructor,refine,obtain,rcases,ext,congr,rfl,trivial,
    noncomputable,axiom,opaque,set_option,attribute},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]",
  literate=
    {→}{{$\to$}}1
    {←}{{$\leftarrow$}}1
    {↔}{{$\leftrightarrow$}}1
    {∀}{{$\forall$}}1
    {∃}{{$\exists$}}1
    {λ}{{$\lambda$}}1
    {≤}{{$\leq$}}1
    {≥}{{$\geq$}}1
    {≠}{{$\neq$}}1
    {∧}{{$\wedge$}}1
    {∨}{{$\vee$}}1
    {¬}{{$\neg$}}1
    {∘}{{$\circ$}}1
    {⊆}{{$\subseteq$}}1
    {ℕ}{{$\mathbb{N}$}}1
    {ℤ}{{$\mathbb{Z}$}}1
    {ℝ}{{$\mathbb{R}$}}1
}

\lstset{
  language=Lean4,
  basicstyle=\small\ttfamily,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{red!70!black},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  numbers=left,
  numberstyle=\tiny\color{gray},
  xleftmargin=2em,
  framexleftmargin=1.5em,
  captionpos=b,
  aboveskip=1em,
  belowskip=1em
}

% ── Headers ───────────────────────────────────────────────────────────
\pagestyle{fancy}
\fancyhf{}
\fancyhead[LE]{\small\itshape\nouppercase{\leftmark}}
\fancyhead[RO]{\small\itshape\nouppercase{\rightmark}}
\fancyfoot[C]{\thepage}
\renewcommand{\headrulewidth}{0.4pt}

% ── Custom commands ───────────────────────────────────────────────────
\newcommand{\IS}{\mathcal{I}}          % Ideatic space
\newcommand{\compose}{\circ}           % Composition operator
\newcommand{\void}{\mathbf{0}}         % Void element
\newcommand{\depth}{\operatorname{d}}  % Depth function
\newcommand{\res}{\sim}                % Resonance relation
\newcommand{\interp}[2]{#1 \compose #2}  % Interpretation
\newcommand{\Rep}{\mathcal{R}}         % Repertoire
\newcommand{\Sig}{\mathcal{S}}         % Signal space
\newcommand{\Pay}{\pi}                 % Payoff

% ── Title ─────────────────────────────────────────────────────────────
\title{
  \vspace{-2cm}
  {\LARGE\textsc{Signalling Games in}}\\[0.3cm]
  {\Huge\textbf{Ideatic Diffusion Theory}}\\[0.5cm]
  {\large\textit{A Formal Theory of Meaning, Interpretation, and Strategic Communication}}\\[1cm]
  {\normalsize With 364 Machine-Verified Theorems in Lean 4}
}
\author{
  \textsc{Halley Young}\\[0.3cm]
  \small\textit{University of Pennsylvania}
}
\date{\today}

% ══════════════════════════════════════════════════════════════════════
\begin{document}

\frontmatter
\maketitle

% ── Epigraph ──────────────────────────────────────────────────────────
\newpage
\thispagestyle{empty}
\vspace*{3cm}
\begin{center}
\textit{``The limits of my language mean the limits of my world.''}\\[0.3cm]
--- Ludwig Wittgenstein, \textit{Tractatus Logico-Philosophicus} (5.6)
\vspace{2cm}

\textit{``A sign is something which stands to somebody for something\\
in some respect or capacity.''}\\[0.3cm]
--- Charles Sanders Peirce, \textit{Collected Papers} (2.228)
\vspace{2cm}

\textit{``Culture is public because meaning is.''}\\[0.3cm]
--- Clifford Geertz, \textit{The Interpretation of Cultures}
\end{center}

% ── Preface ───────────────────────────────────────────────────────────
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book develops a formal theory of signalling games within the framework
of Ideatic Diffusion Theory (IDT). Where classical game theory treats
signals as tokens in an abstract communication channel, we treat them
as \emph{ideas}---elements of an ideatic space equipped with composition,
depth, and resonance. The central insight is that \emph{interpretation
is composition}: when a receiver with background idea $r$ encounters
signal $s$, they do not simply ``decode'' it but rather \emph{compose}
it with their existing understanding, producing the interpretation
$r \compose s$.

This compositional view of interpretation has profound consequences for
game theory. Payoffs depend not on the signal itself but on what
receivers \emph{make of it}---how it composes with their existing repertoire
of ideas. A signal that is profound to one receiver may be trivial to
another, not because the signal differs but because their repertoires do.
This is the formal kernel of hermeneutic theory: meaning is not in the
text but in the encounter between text and reader.

The book is organized in four parts. Part~I (\Cref{ch:hermeneutic,ch:signalling,ch:costly,ch:activation,ch:misinterp})
establishes the interpretation framework: repertoires, signalling problems,
costly signalling, resonance activation, and misinterpretation. Part~II
(\Cref{ch:equilibrium,ch:persuasion,ch:lossy,ch:echo,ch:multireceiver})
develops equilibrium theory, persuasion bounds, information loss, echo
chambers, and multi-receiver signalling. Part~III
(\Cref{ch:marketplace,ch:matching,ch:coalition,ch:auction,ch:curation})
studies populations: competition, matching, coalitions, auctions, and
repertoire curation. Part~IV
(\Cref{ch:cooperative,ch:bargaining,ch:network,ch:social,ch:synthesis})
addresses cooperation, bargaining, network diffusion, social choice,
and offers a synthesis.

Every theorem in this book has been formally verified in Lean~4,
building on the \texttt{IDT\_Foundations} library. There are no
\texttt{sorry} axioms. The 364 machine-checked theorems constitute,
to our knowledge, the first comprehensive formal treatment of signalling
games in an algebraic theory of ideas.

\vspace{1cm}
\noindent\textit{Philadelphia, 2026}

% ── Table of Contents ─────────────────────────────────────────────────
\tableofcontents

% ══════════════════════════════════════════════════════════════════════
\mainmatter

% ── Part I: The Interpretation Framework ──────────────────────────────
\part{The Interpretation Framework}
\label{part:interpretation}

\input{tex_chapters/ch01}
\input{tex_chapters/ch02}
\input{tex_chapters/ch03}
\input{tex_chapters/ch04}
\input{tex_chapters/ch05}

% ── Part II: Equilibrium and Dynamics ─────────────────────────────────
\part{Equilibrium and Dynamics}
\label{part:equilibrium}

\input{tex_chapters/ch06}
\input{tex_chapters/ch07}
\input{tex_chapters/ch08}
\input{tex_chapters/ch09}
\input{tex_chapters/ch10}

% ── Part III: Populations and Markets ─────────────────────────────────
\part{Populations and Markets}
\label{part:populations}

\input{tex_chapters/ch11}
\input{tex_chapters/ch12}
\input{tex_chapters/ch13}
\input{tex_chapters/ch14}
\input{tex_chapters/ch15}

% ── Part IV: Cooperation, Networks, and Synthesis ─────────────────────
\part{Cooperation, Networks, and Synthesis}
\label{part:synthesis}

\input{tex_chapters/ch16}
\input{tex_chapters/ch17}
\input{tex_chapters/ch18}
\input{tex_chapters/ch19}
\input{tex_chapters/ch20}

% ══════════════════════════════════════════════════════════════════════
\backmatter

% ── Bibliography ──────────────────────────────────────────────────────
\begin{thebibliography}{99}

\bibitem{austin1962} Austin, J.L. (1962). \textit{How to Do Things with Words}. Oxford University Press.

\bibitem{bartlett1932} Bartlett, F.C. (1932). \textit{Remembering: A Study in Experimental and Social Psychology}. Cambridge University Press.

\bibitem{bateson1972} Bateson, G. (1972). \textit{Steps to an Ecology of Mind}. University of Chicago Press.

\bibitem{bourdieu1984} Bourdieu, P. (1984). \textit{Distinction: A Social Critique of the Judgement of Taste}. Harvard University Press.

\bibitem{bourdieu1991} Bourdieu, P. (1991). \textit{Language and Symbolic Power}. Harvard University Press.

\bibitem{crawford1982} Crawford, V.P. \& Sobel, J. (1982). Strategic information transmission. \textit{Econometrica}, 50(6), 1431--1451.

\bibitem{dawkins1976} Dawkins, R. (1976). \textit{The Selfish Gene}. Oxford University Press.

\bibitem{derrida1967} Derrida, J. (1967). \textit{Of Grammatology}. Johns Hopkins University Press.

\bibitem{durkheim1893} Durkheim, \'E. (1893). \textit{The Division of Labor in Society}. Free Press.

\bibitem{durkheim1912} Durkheim, \'E. (1912). \textit{The Elementary Forms of Religious Life}. Free Press.

\bibitem{eco1976} Eco, U. (1976). \textit{A Theory of Semiotics}. Indiana University Press.

\bibitem{foucault1972} Foucault, M. (1972). \textit{The Archaeology of Knowledge}. Pantheon Books.

\bibitem{gadamer1960} Gadamer, H.-G. (1960). \textit{Truth and Method}. Continuum.

\bibitem{geertz1973} Geertz, C. (1973). \textit{The Interpretation of Cultures}. Basic Books.

\bibitem{goffman1959} Goffman, E. (1959). \textit{The Presentation of Self in Everyday Life}. Anchor Books.

\bibitem{gramsci1971} Gramsci, A. (1971). \textit{Selections from the Prison Notebooks}. International Publishers.

\bibitem{habermas1984} Habermas, J. (1984). \textit{The Theory of Communicative Action}. Beacon Press.

\bibitem{keats1817} Keats, J. (1817). Letter to George and Tom Keats, 21 December 1817.

\bibitem{levistrauss1962} L\'evi-Strauss, C. (1962). \textit{The Savage Mind}. University of Chicago Press.

\bibitem{levistrauss1969} L\'evi-Strauss, C. (1969). \textit{The Elementary Structures of Kinship}. Beacon Press.

\bibitem{lewis1969} Lewis, D. (1969). \textit{Convention: A Philosophical Study}. Harvard University Press.

\bibitem{mauss1925} Mauss, M. (1925). \textit{The Gift}. Norton.

\bibitem{nash1950} Nash, J. (1950). Equilibrium points in $n$-person games. \textit{PNAS}, 36(1), 48--49.

\bibitem{peirce1931} Peirce, C.S. (1931--1958). \textit{Collected Papers}. Harvard University Press.

\bibitem{rappaport1999} Rappaport, R.A. (1999). \textit{Ritual and Religion in the Making of Humanity}. Cambridge University Press.

\bibitem{ricoeur1976} Ricoeur, P. (1976). \textit{Interpretation Theory: Discourse and the Surplus of Meaning}. Texas Christian University Press.

\bibitem{saussure1916} Saussure, F. de (1916). \textit{Course in General Linguistics}. Open Court.

\bibitem{schelling1960} Schelling, T.C. (1960). \textit{The Strategy of Conflict}. Harvard University Press.

\bibitem{searle1969} Searle, J.R. (1969). \textit{Speech Acts}. Cambridge University Press.

\bibitem{simmel1908} Simmel, G. (1908). \textit{Soziologie}. Duncker \& Humblot.

\bibitem{sosis2003} Sosis, R. (2003). Why aren't we all Hutterites? Costly signaling theory and religious behavior. \textit{Human Nature}, 14(2), 91--127.

\bibitem{spence1973} Spence, M. (1973). Job market signaling. \textit{QJE}, 87(3), 355--374.

\bibitem{sperber1996} Sperber, D. (1996). \textit{Explaining Culture: A Naturalistic Approach}. Blackwell.

\bibitem{sperberw1986} Sperber, D. \& Wilson, D. (1986). \textit{Relevance: Communication and Cognition}. Blackwell.

\bibitem{turner1969} Turner, V. (1969). \textit{The Ritual Process: Structure and Anti-Structure}. Aldine.

\bibitem{veblen1899} Veblen, T. (1899). \textit{The Theory of the Leisure Class}. Macmillan.

\bibitem{vygotsky1978} Vygotsky, L.S. (1978). \textit{Mind in Society}. Harvard University Press.

\bibitem{wittgenstein1953} Wittgenstein, L. (1953). \textit{Philosophical Investigations}. Blackwell.

\bibitem{zahavi1975} Zahavi, A. (1975). Mate selection---a selection for a handicap. \textit{Journal of Theoretical Biology}, 53(1), 205--214.

\end{thebibliography}

% ── Index placeholder ─────────────────────────────────────────────────
\chapter*{Index of Theorems}
\addcontentsline{toc}{chapter}{Index of Theorems}

The following theorems are formally verified in Lean~4 with zero
\texttt{sorry} axioms. Each theorem name links to the corresponding
Lean source file in the \texttt{FormalAnthropology/} directory.

\begin{small}
\begin{enumerate}
\item[\textbf{Ch.~1}] \texttt{total\_hermeneutic\_bound}, \texttt{max\_interp\_depth\_bound}, \texttt{silence\_is\_transparent}, \texttt{blank\_receiver}, \texttt{negative\_capability}, \texttt{durkheimian\_consensus}, \texttt{mythical\_transformation}, \texttt{turners\_communitas}, \texttt{bartletts\_serial\_reproduction}, \texttt{telephone\_depth}, \texttt{resonant\_transmission\_lines}, \texttt{void\_is\_unpersuasive}, \texttt{interpretation\_conserves\_count}, \texttt{durkheims\_division\_of\_labor}, \texttt{silence\_preserves\_depthSum}, \texttt{three\_stage\_telephone}

\item[\textbf{Ch.~2--5}] See \texttt{IDT\_Signal\_Ch02--05.lean}
\item[\textbf{Ch.~6--10}] See \texttt{IDT\_Signal\_Ch06--10.lean}
\item[\textbf{Ch.~11--15}] See \texttt{IDT\_Signal\_Ch11--15.lean}
\item[\textbf{Ch.~16--20}] See \texttt{IDT\_Signal\_Ch16--20.lean}
\end{enumerate}
\end{small}

\end{document}
