\documentclass[11pt,oneside]{book}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usepackage{epigraph}
\usepackage{tocloft}

% Theorem environments
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{axiom_def}[theorem]{Axiom}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{interpretation}[theorem]{Interpretation}

% Lean code listings
\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,noncomputable,class,instance,where,
    variable,open,namespace,section,end,import,structure,inductive,
    match,with,fun,let,in,if,then,else,do,return,have,show,by,
    exact,rfl,simp,linarith,nlinarith,ring,intro,apply,rw,unfold,
    calc,sorry,Prop,Type,Sort,axiom,extends,deriving},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]",
}

\lstset{
  language=Lean4,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{red},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  xleftmargin=1em,
  xrightmargin=1em,
}

% Custom commands
\newcommand{\rs}{\mathrm{rs}}
\newcommand{\emerge}{\kappa}
\newcommand{\compose}{\circ}
\newcommand{\void}{\varepsilon}
\newcommand{\IS}{\mathcal{I}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}

\title{\Huge\bfseries The Math of Ideas\\[0.3em]
\LARGE Volume I: Foundations\\[0.5em]
\Large Composition, Resonance, and Emergence from Eight Axioms}
\author{A Formal Treatment with Machine-Verified Proofs\\[0.3em]
\normalsize\textit{Part of the series ``The Math of Ideas''}}
\date{}

\begin{document}

\maketitle

\frontmatter

\chapter*{Preface}

This is the first volume in a six-volume series, \emph{The Math of Ideas}, which 
develops a unified mathematical framework for studying ideas---how they compose, 
resonate, and give rise to genuinely new meaning. This volume establishes the 
foundations: eight axioms from which all subsequent volumes build. Unlike existing 
approaches to memetics, cultural evolution, or information theory, the math of 
ideas takes the \emph{idea} as a primitive object and derives its properties from 
a minimal set of axioms about composition and resonance.

The five volumes that follow study the same mathematical universe from different 
vantage points: strategic interaction (Vol.~II), philosophy of meaning (Vol.~III), 
signs and culture (Vol.~IV), power and ethics (Vol.~V), and emergence itself 
(Vol.~VI). But they all rest on the theorems proven here. This volume is therefore 
not merely ``pure math before the applications''---it is the common language in 
which all questions about ideas can be posed and, crucially, \emph{proven}.

Every major theorem in this volume is interpreted through three lenses---literary 
theory, philosophy, and semiotics---because the mathematics of ideas is not an 
analogy \emph{applied to} the humanities; it \emph{is} the humanities, written in 
the only language precise enough to prove things.

The central mathematical object is the \emph{ideatic space} $(\IS, \compose, \void, \rs)$:
a monoid of ideas equipped with a real-valued resonance function. The key innovation
is the \emph{emergence term}
\[
\emerge(a,b,c) := \rs(a \compose b, c) - \rs(a,c) - \rs(b,c),
\]
which captures the genuinely new meaning that arises when ideas combine---meaning
that is irreducible to the parts. When $\emerge = 0$, we recover a linear (Hilbert-like)
theory; when $\emerge \neq 0$, we enter the domain of genuine creativity, metaphor,
irony, and dialectical synthesis.

Every theorem in this book has been formally verified in Lean 4 with zero \texttt{sorry}
axioms. The Lean source files are referenced throughout. This is not merely a
guarantee of correctness; it is a demonstration that the theory of ideas admits
the same standard of rigor as algebra or analysis.

\vspace{1em}

The theory proceeds from just eight axioms:
\begin{enumerate}
\item Three \textbf{monoid axioms}: ideas compose associatively with a neutral element (void/silence).
\item Two \textbf{void-resonance axioms}: silence resonates with nothing.
\item Four \textbf{emergence axioms}: self-coherence is non-negative, only silence has zero presence,
emergence satisfies a Cauchy--Schwarz bound, and composition enriches.
\end{enumerate}

From these eight axioms, we derive a remarkably rich theory encompassing:
semiotics (the study of signs), hermeneutics (interpretation theory),
dialectics (thesis--antithesis--synthesis), rhetoric (persuasion),
narrative theory, game theory, evolutionary dynamics, translation theory,
and the geometry of meaning.

\tableofcontents

\mainmatter

%==========================================================================
\part{Foundations}
%==========================================================================

\input{tex_v2_chapters/idt_ch1_7}

%==========================================================================
\part{Structure and Dynamics}
%==========================================================================

\input{tex_v2_chapters/idt_ch8_14}

%==========================================================================
\part{Applications and Synthesis}
%==========================================================================

\input{tex_v2_chapters/idt_ch15_22}

%==========================================================================
% END MATTER
%==========================================================================

\chapter*{Appendix: Complete Lean 4 Formalization}

All theorems in this book have been formally verified in Lean 4 using the
\texttt{IdeaticSpace} type class defined in \texttt{IDT\_v3\_Foundations.lean}.
The complete formalization comprises the following files:

\begin{itemize}
\item \texttt{IDT\_v3\_Foundations.lean} --- 103 theorems. The axiom class,
emergence definition, cocycle condition, self-resonance, semiotics, models.

\item \texttt{IDT\_v3\_Emergence.lean} --- 199 theorems. Cocycle algebra,
higher cocycles, spectrum classification, curvature, stability, differential
calculus, chain algebra, convexity.

\item \texttt{IDT\_v3\_Interpretation.lean} --- 230 theorems. Gadamer,
Ricoeur, Schleiermacher, Benjamin, Heidegger, Derrida, hermeneutic circle,
fusion of horizons, tradition, intertextuality.

\item \texttt{IDT\_v3\_SignalGames.lean} --- 208 theorems. Nash equilibrium,
babbling equilibria, costly signaling, revelation, strategic silence,
signaling chains, grand unification.

\item \texttt{IDT\_v3\_Wittgenstein.lean} --- 183 theorems. Language games,
rule-following, private language, family resemblance, hinge propositions,
Tractatus vs.\ Investigations, showing vs.\ saying, philosophical therapy.

\item \texttt{IDT\_v3\_Semiotics.lean} --- 158 theorems. Saussure, Peirce,
Barthes, Eco's Open Work, Derrida's diff\'erance, Jakobson's six functions,
Greimas' semiotic square.

\item \texttt{IDT\_v3\_PowerStructure.lean} --- 152 theorems. Bourdieu,
Gramsci, Foucault, Althusser, hegemony, ideology, propaganda, censorship,
resistance, liberation dynamics.

\item \texttt{IDT\_v3\_Networks.lean} --- 120 theorems. Network topology,
diffusion dynamics, information cascades, community structure, influence.

\item \texttt{IDT\_v3\_Poetics.lean} --- 103 theorems. Condensation,
defamiliarization, meter, rhyme, irony, metaphor economy.

\item \texttt{IDT\_v3\_Phenomenology.lean} --- 90 theorems. Husserl,
Heidegger, Merleau-Ponty, intentionality, lifeworld, embodiment.

\item \texttt{IDT\_v3\_DeepGames.lean} --- 72 theorems. Transparency,
cocycle algebra, order sensitivity, attribution impossibility, weight ratchet,
coalitions, fundamental synthesis.

\item \texttt{IDT\_v3\_Diffusion.lean} --- 38 theorems. Transmission,
chains, mutation, conservative/mutagenic receivers.

\item \texttt{IDT\_v3\_Geometry.lean} --- 40 theorems. Distances,
orthogonality, curvature, projections.

\item \texttt{IDT\_v3\_Narrative.lean} --- 40 theorems. Plot, defamiliarization,
narrative arcs, rhetoric.

\item \texttt{IDT\_v3\_Cooperation.lean} --- 26 theorems. Coalition value,
cooperation surplus, Shapley values.

\item \texttt{IDT\_v3\_Evolution.lean} --- 24 theorems. Fitness, invasion,
Red Queen, memetic drift.

\item \texttt{IDT\_v3\_Translation.lean} --- 21 theorems. Faithful translation,
untranslatability, round-trip fidelity.

\item \texttt{IDT\_v3\_Dialectics.lean} --- 20 theorems. Opposition,
synthesis, Aufhebung, iterated dialectics.
\end{itemize}

\noindent\textbf{Total: 1,827 theorems across 18 files, all compiling with zero
\texttt{sorry} axioms.}

\chapter*{Bibliography}

\begin{enumerate}[label={[\arabic*]}]
\item Aristotle, \emph{Metaphysics}.
\item Austin, J.L., \emph{How to Do Things with Words}, Oxford, 1962.
\item Bakhtin, M.M., \emph{The Dialogic Imagination}, University of Texas Press, 1981.
\item Barthes, R., \emph{S/Z}, Hill and Wang, 1974.
\item Bloom, H., \emph{The Anxiety of Influence}, Oxford, 1973.
\item Dawkins, R., \emph{The Selfish Gene}, Oxford, 1976.
\item Derrida, J., \emph{Of Grammatology}, Johns Hopkins, 1976.
\item Gadamer, H.-G., \emph{Truth and Method}, Continuum, 1975.
\item Grice, H.P., ``Logic and Conversation,'' in \emph{Syntax and Semantics 3}, 1975.
\item Hegel, G.W.F., \emph{Phenomenology of Spirit}, Oxford, 1807/1977.
\item Kristeva, J., ``Word, Dialogue, and Novel,'' in \emph{Desire in Language}, Columbia, 1980.
\item Lewis, D., \emph{Convention}, Harvard, 1969.
\item Nash, J., ``Non-Cooperative Games,'' \emph{Annals of Mathematics} 54(2), 1951.
\item Peirce, C.S., \emph{Collected Papers}, Harvard, 1931--1958.
\item de Saussure, F., \emph{Course in General Linguistics}, Open Court, 1916/1986.
\item Shannon, C.E., ``A Mathematical Theory of Communication,'' \emph{Bell System Technical Journal}, 1948.
\item Shklovsky, V., ``Art as Technique,'' 1917.
\item von Neumann, J.\ and Morgenstern, O., \emph{Theory of Games and Economic Behavior}, Princeton, 1944.
\item Wittgenstein, L., \emph{Philosophical Investigations}, Blackwell, 1953.
\item Wittgenstein, L., \emph{Tractatus Logico-Philosophicus}, Routledge, 1921/1922.
\end{enumerate}

\end{document}
