% ================================================================
% IDEATIC DIFFUSION THEORY v2:
% The Hilbert Space of Meaning
% ================================================================
\documentclass[11pt,openany]{book}

% --- Geometry ---
\usepackage[margin=1in]{geometry}

% --- Fonts and encoding ---
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}

% --- Math ---
\usepackage{amsmath,amssymb,amsthm,mathtools}

% --- Graphics and colors ---
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}

% --- Cross-referencing ---
\usepackage[colorlinks=true,linkcolor=MidnightBlue,citecolor=OliveGreen,urlcolor=BrickRed]{hyperref}
\usepackage{cleveref}

% --- Code listings for Lean ---
\usepackage{listings}
\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,class,structure,instance,where,
    variable,import,open,namespace,end,section,
    by,exact,intro,induction,cases,rfl,simp,omega,calc,have,show,
    apply,rw,push_neg,with,generalizing,let,match,if,then,else,
    return,do,for,in,private,noncomputable,set_option},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[n]{/-}{-/},
  morestring=[b]",
  literate={→}{{$\to$}}1 {←}{{$\leftarrow$}}1 {↦}{{$\mapsto$}}1
           {∀}{{$\forall$}}1 {∃}{{$\exists$}}1 {λ}{{$\lambda$}}1
           {≤}{{$\leq$}}1 {≥}{{$\geq$}}1 {≠}{{$\neq$}}1
           {∧}{{$\wedge$}}1 {∨}{{$\vee$}}1 {¬}{{$\neg$}}1
           {∈}{{$\in$}}1 {∉}{{$\notin$}}1 {⊆}{{$\subseteq$}}1
           {ℕ}{{$\mathbb{N}$}}1 {ℤ}{{$\mathbb{Z}$}}1 {ℝ}{{$\mathbb{R}$}}1
           {∘}{{$\circ$}}1 {×}{{$\times$}}1
}
\lstset{
  language=Lean4,
  basicstyle=\small\ttfamily,
  keywordstyle=\color{MidnightBlue}\bfseries,
  commentstyle=\color{OliveGreen}\itshape,
  stringstyle=\color{BrickRed},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  numbers=none,
  xleftmargin=1em,
  xrightmargin=1em,
  aboveskip=1em,
  belowskip=1em,
}

% --- Epigraphs ---
\usepackage{epigraph}
\setlength{\epigraphwidth}{0.7\textwidth}

% --- Theorem environments ---
\theoremstyle{definition}
\newtheorem{axiom}{Axiom}[chapter]
\newtheorem{definition}[axiom]{Definition}

\theoremstyle{plain}
\newtheorem{theorem}[axiom]{Theorem}
\newtheorem{proposition}[axiom]{Proposition}
\newtheorem{lemma}[axiom]{Lemma}
\newtheorem{corollary}[axiom]{Corollary}

\theoremstyle{remark}
\newtheorem{remark}[axiom]{Remark}
\newtheorem{example}[axiom]{Example}
\newtheorem{interpretation}[axiom]{Interpretation}

% --- Custom commands (v1 compatibility) ---
\newcommand{\IS}{\mathcal{I}}               % Ideatic space
\newcommand{\compose}{\circ}                 % Composition
\newcommand{\void}{\mathbf{0}}              % Void
\newcommand{\res}{\sim}                      % Resonance
\newcommand{\depth}{\operatorname{d}}       % Depth
\newcommand{\transmit}{\tau}                 % Transmission
\newcommand{\compN}[2]{#1^{\langle #2 \rangle}} % composeN
\newcommand{\Filt}[1]{\mathcal{F}_{#1}}     % Depth filtration
\newcommand{\orbit}[2]{#1^{[#2]}}           % Iterate/orbit

% --- Custom commands (v2: Hilbert space) ---
\newcommand{\Hspace}{\mathcal{H}}            % Hilbert space
\newcommand{\embed}[1]{\varphi(#1)}          % Embedding φ(a)
\newcommand{\rs}[2]{\langle\!\langle #1, #2 \rangle\!\rangle}  % resonanceStrength
\newcommand{\ip}[2]{\langle #1, #2 \rangle}  % inner product
\newcommand{\nrm}[1]{\lVert #1 \rVert}       % norm
\newcommand{\proj}[2]{\operatorname{proj}_{#1}\!\left(#2\right)} % projection
\newcommand{\cosangle}[2]{\theta(#1,#2)}     % angle between ideas
\newcommand{\Gram}{\mathbf{G}}               % Gram matrix
\newcommand{\RMat}{\mathbf{R}}               % Resonance matrix
\newcommand{\payoff}[2]{\pi_{#1}(#2)}        % payoff

\newcommand{\lean}[1]{\lstinline[language=Lean4]{#1}}

% --- Title ---
\title{%
  \vspace{-2cm}
  {\LARGE\scshape Ideatic Diffusion Theory v2}\\[0.5cm]
  {\Large The Hilbert Space of Meaning}\\[1cm]
  {\large A Mathematical Theory of Ideas, Resonance,\\
  and Strategic Interaction,\\[0.3cm]
  with Machine-Verified Proofs in Lean~4}
}
\author{}
\date{}

\begin{document}

% ================================================================
% FRONT MATTER
% ================================================================
\frontmatter

\maketitle
\thispagestyle{empty}

\vfill
\begin{center}
\textit{For every mind that has ever felt\\
the geometry of a thought taking shape.}
\end{center}
\vfill

\clearpage

% --- Abstract ---
\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

This book develops \textbf{Ideatic Diffusion Theory v2} (IDT\,v2), a mathematical
framework in which ideas are embedded into a real Hilbert space $\Hspace$ via
a structure-preserving map $\varphi : \IS \to \Hspace$. The central innovation
is that \emph{resonance} between ideas becomes an \emph{inner product}:
\[
  \rs{a}{b} \;=\; \ip{\embed{a}}{\embed{b}},
\]
and \emph{composition} of ideas maps to vector \emph{addition}:
\[
  \embed{a \compose b} \;=\; \embed{a} + \embed{b}.
\]
This simple pair of axioms yields a remarkably rich geometric theory of meaning.
The Cauchy--Schwarz inequality becomes a fundamental limit on cross-cultural
understanding. The distance between ideas becomes a genuine metric---making
Kuhn's ``incommensurability'' measurable rather than metaphorical. Persuasion
reduces to orthogonal projection, Nash equilibria in meaning games admit
geometric characterization, and the spectral decomposition of the resonance
matrix reveals the principal modes of cultural agreement.

The book is organized in four parts. Part~I (Chapters~\ref{ch:geometry}--\ref{ch:orthogonality})
establishes the Hilbert space foundations: the embedding axioms, resonance as
inner product, the metric of meaning, the relationship between depth and norm,
and the structure of orthogonal (unrelated) ideas. Part~II
(Chapters~\ref{ch:payoffs}--\ref{ch:spectral}) develops the game theory of
meaning: payoffs as inner products, Nash equilibria in Hilbert space,
cooperation and conflict as angular relationships, persuasion as projection,
and spectral theory of cultural networks. Parts~III and~IV (Chapters~11--20)
extend the theory to dynamics, learning, and applications across the humanities.

Every theorem has been verified by the Lean~4 proof assistant, ensuring absolute
logical correctness. The formalization demonstrates that the Hilbert space
embedding is not merely a metaphor but a precise mathematical structure with
non-trivial consequences.

\clearpage

% --- Preface ---
\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\epigraph{The limits of my language mean the limits of my world.}{Ludwig Wittgenstein, \textit{Tractatus Logico-Philosophicus}}

The first edition of this book---\textit{Ideatic Diffusion Theory: A
Mathematical Foundation for the Study of Ideas}---introduced ideas as elements
of a monoid with a reflexive, symmetric, non-transitive resonance relation and
a subadditive depth function. That framework yielded rich structural theorems
about composition, transmission, and cultural dynamics, but it left a
fundamental question unanswered: \emph{how strong} is the resonance between two
ideas?

In IDT~v1, resonance was a binary relation: two ideas either resonate or they
do not. This is a useful abstraction, but it misses the quantitative nuance that
pervades actual intellectual life. Darwin's theory of natural selection resonates
more strongly with Wallace's independent discovery than with Lamarck's earlier
theory of inheritance---but all three resonate to some degree. Binary resonance
cannot express this gradation.

IDT~v2 resolves this limitation by embedding the ideatic space into a real
Hilbert space and identifying resonance strength with the inner product. The
result is a \emph{geometry of meaning}: ideas become vectors, resonance becomes
angle, composition becomes addition, and the full apparatus of functional
analysis becomes available for studying the structure of thought.

This is not merely a technical refinement. The passage from binary resonance to
inner-product resonance is analogous to the passage from qualitative to
quantitative physics---from ``heavy'' and ``light'' to mass measured in
kilograms. It transforms IDT from a qualitative structural theory into a
quantitative geometric one, opening entirely new domains of application:
game-theoretic analysis of communication, spectral analysis of cultural
networks, optimization-theoretic accounts of persuasion, and metric-space
approaches to intellectual history.

\subsection*{How to Read This Book}

This book addresses three audiences simultaneously:

\begin{itemize}
\item \textbf{For humanists}: Every mathematical result is accompanied by an
``Interpretation'' block explaining its significance for philosophy, literary
theory, anthropology, or cultural studies. You may skip proofs on first reading.

\item \textbf{For mathematicians}: The Hilbert space framework is standard, but
the interpretive lens is novel. The game-theoretic and spectral-theoretic
results have genuine mathematical content.

\item \textbf{For formalists}: Every theorem is machine-verified in Lean~4. The
complete formalization is available as a companion repository.
\end{itemize}

\subsection*{What Is New in v2}

The key innovations of this second edition are:

\begin{enumerate}
\item \textbf{Hilbert space embedding}: Ideas map to vectors in a real Hilbert
space $\Hspace$ via $\varphi : \IS \to \Hspace$.

\item \textbf{Quantitative resonance}: $\rs{a}{b} = \ip{\embed{a}}{\embed{b}}$
replaces the binary resonance relation.

\item \textbf{Composition as addition}: $\embed{a \compose b} = \embed{a} +
\embed{b}$ gives composition geometric content.

\item \textbf{Game theory of meaning}: Communication games where payoffs are
inner products, with geometric characterization of equilibria.

\item \textbf{Spectral theory of culture}: Eigendecomposition of the resonance
(Gram) matrix reveals principal cultural dimensions.

\item \textbf{Persuasion as projection}: The geometry of how ideas influence
each other, formalized as orthogonal projection.
\end{enumerate}

\subsection*{Acknowledgments}

The development of IDT~v2 owes debts to many traditions: to Peter G\"ardenfors,
whose \emph{Conceptual Spaces} anticipated the idea of embedding meanings in
geometric spaces; to Charles Osgood, whose semantic differential pioneered the
measurement of meaning; to Tomas Mikolov and colleagues, whose word2vec
demonstrated that distributional semantics yields genuine geometric structure;
and to the communities of Lean~4 and Mathlib, whose tools made machine
verification possible.

\clearpage

% --- Table of Contents ---
\setcounter{tocdepth}{2}
\tableofcontents

\clearpage

% ================================================================
% MAIN MATTER
% ================================================================
\mainmatter

% ================================================================
% PART I: FOUNDATIONS OF THE HILBERT THEORY
% ================================================================
\part{Foundations of the Hilbert Theory}
\label{part:foundations}

\input{tex_v2_chapters/ch01}
\input{tex_v2_chapters/ch02}
\input{tex_v2_chapters/ch03}
\input{tex_v2_chapters/ch04}
\input{tex_v2_chapters/ch05}

% ================================================================
% PART II: THE GAME THEORY OF MEANING
% ================================================================
\part{The Game Theory of Meaning}
\label{part:game-theory}

\input{tex_v2_chapters/ch06}
\input{tex_v2_chapters/ch07}
\input{tex_v2_chapters/ch08}
\input{tex_v2_chapters/ch09}
\input{tex_v2_chapters/ch10}

% ================================================================
% PART III: DYNAMICS AND LEARNING IN MEANING SPACE
% ================================================================
\part{Dynamics and Learning in Meaning Space}
\label{part:dynamics}

\input{tex_v2_chapters/ch11}
\input{tex_v2_chapters/ch12}
\input{tex_v2_chapters/ch13}
\input{tex_v2_chapters/ch14}
\input{tex_v2_chapters/ch15}

% ================================================================
% PART IV: APPLICATIONS ACROSS THE HUMANITIES
% ================================================================
\part{Applications Across the Humanities}
\label{part:applications}

\input{tex_v2_chapters/ch16}
\input{tex_v2_chapters/ch17}
\input{tex_v2_chapters/ch18}
\input{tex_v2_chapters/ch19}
\input{tex_v2_chapters/ch20}

% ================================================================
% BACK MATTER
% ================================================================
\backmatter

\chapter*{Notation Index}
\addcontentsline{toc}{chapter}{Notation Index}

\begin{tabular}{lll}
\textbf{Symbol} & \textbf{Meaning} & \textbf{Introduced} \\
\hline
$\IS$ & Ideatic space & Ch.~\ref{ch:geometry} \\
$\Hspace$ & Hilbert space of meanings & Ch.~\ref{ch:geometry} \\
$\varphi$ & Embedding $\IS \to \Hspace$ & Ch.~\ref{ch:geometry} \\
$\compose$ & Composition of ideas & Ch.~\ref{ch:geometry} \\
$\void$ & Void (identity/zero vector) & Ch.~\ref{ch:geometry} \\
$\rs{a}{b}$ & Resonance strength & Ch.~\ref{ch:resonance} \\
$\ip{x}{y}$ & Inner product in $\Hspace$ & Ch.~\ref{ch:resonance} \\
$\nrm{x}$ & Norm in $\Hspace$ & Ch.~\ref{ch:resonance} \\
$d(a,b)$ & Meaning distance & Ch.~\ref{ch:metric} \\
$\depth(a)$ & Depth (combinatorial complexity) & Ch.~\ref{ch:depth-norm} \\
$\cosangle{a}{b}$ & Angle between ideas & Ch.~\ref{ch:cooperation} \\
$\proj{v}{w}$ & Projection of $w$ onto $v$ & Ch.~\ref{ch:persuasion} \\
$\RMat$ & Resonance matrix & Ch.~\ref{ch:spectral} \\
$\Gram$ & Gram matrix & Ch.~\ref{ch:spectral} \\
$\payoff{i}{s}$ & Payoff of player $i$ & Ch.~\ref{ch:payoffs} \\
\end{tabular}

\end{document}
