\documentclass[11pt,oneside]{book}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usepackage{epigraph}

% Theorem environments
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{axiom_def}[theorem]{Axiom}
\newtheorem{game}[theorem]{Game}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{interpretation}[theorem]{Interpretation}
\newtheorem{aside}[theorem]{Aside}

% Lean code listings
\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,noncomputable,class,instance,where,
    variable,open,namespace,section,end,import,structure,inductive,
    match,with,fun,let,in,if,then,else,do,return,have,show,by,
    exact,rfl,simp,linarith,nlinarith,ring,intro,apply,rw,unfold,
    calc,sorry,Prop,Type,Sort,axiom,extends,deriving},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]",
}

\lstset{
  language=Lean4,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{red},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  xleftmargin=1em,
  xrightmargin=1em,
}

% Custom commands
\newcommand{\rs}{\mathrm{rs}}
\newcommand{\emerge}{\kappa}
\newcommand{\compose}{\circ}
\newcommand{\void}{\varepsilon}
\newcommand{\IS}{\mathcal{I}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}

\title{\Huge\bfseries The Math of Ideas\\[0.3em]
\LARGE Volume II: Strategic Interaction\\[0.5em]
\Large Games, Networks, and the Evolution of Meaning}
\author{A Formal Treatment with Machine-Verified Proofs\\[0.3em]
\normalsize\textit{Part of the series ``The Math of Ideas''}}
\date{}

\begin{document}

\maketitle

\frontmatter

\chapter*{Preface: A Reply to Wittgenstein and the Game Theorists}

This book is a reply to two traditions that have, for too long, talked past
each other.

To the \emph{game theorists}: you have built a magnificent edifice of
strategic reasoning, but you treat the objects of exchange as given---goods,
money, utility. When agents exchange \emph{ideas}, the payoff structure is
fundamentally different. The combination of two ideas produces something
genuinely new: an \emph{emergence} that is irreducible to the parts. This
book shows that emergence transforms every classical game-theoretic result:
equilibria shift, cooperation surpluses appear, and order effects create
asymmetries that have no classical counterpart.

To the followers of \emph{Wittgenstein}: you were right that ``meaning is use''
(\emph{Philosophical Investigations} \S43), that language games are the
foundation of meaning, and that the private language argument shows meaning
to be inherently social. But you stopped too soon. This book shows that
Wittgenstein's deepest insights---the beetle in the box, the rule-following
paradox, the impossibility of a private language---are not merely philosophical
arguments but \emph{mathematical theorems}, provable from eight axioms about how
ideas compose and resonate. The beetle \emph{must} drop out of the box. Private
languages \emph{must} collapse to silence. And the reason is emergence.

The predecessor volume, \emph{Idea Diffusion Theory}, established the
mathematical framework: the ideatic space $(\IS, \compose, \void, \rs)$ with
its eight axioms, the emergence term $\emerge(a,b,c) := \rs(a \compose b, c) -
\rs(a,c) - \rs(b,c)$, and a rich theory of composition, transmission, and
interpretation. This sequel assumes those foundations and develops the
game-theoretic consequences.

Every theorem has been verified in Lean 4 with zero \texttt{sorry} axioms.

\tableofcontents

\mainmatter

%==========================================================================
% PROLOGUE: RECALLING THE FOUNDATIONS
%==========================================================================

\chapter*{Prologue: The Ideatic Space in Brief}
\addcontentsline{toc}{chapter}{Prologue: The Ideatic Space in Brief}

\textit{This chapter recalls the essential definitions and results from
\emph{Idea Diffusion Theory} that are needed for the game-theoretic
development. Readers who have read the predecessor volume may skim this
chapter; those who have not should read it carefully.}

\section*{The Eight Axioms}

We work with an \emph{ideatic space} $(\IS, \compose, \void, \rs)$ where
$\IS$ is a set of ideas, $\compose : \IS \times \IS \to \IS$ is composition,
$\void \in \IS$ is the void idea (silence), and $\rs : \IS \times \IS \to \R$
is the \emph{resonance function}. The axioms are:

\begin{enumerate}
\item \textbf{Associativity}: $(a \compose b) \compose c = a \compose (b \compose c)$.
\item \textbf{Left identity}: $\void \compose a = a$.
\item \textbf{Right identity}: $a \compose \void = a$.
\item \textbf{Void resonance (right)}: $\rs(a, \void) = 0$.
\item \textbf{Void resonance (left)}: $\rs(\void, a) = 0$.
\item \textbf{Self-coherence}: $\rs(a, a) \geq 0$, with $\rs(a,a) = 0 \iff a = \void$.
\item \textbf{Emergence bound}: $\emerge(a,b,c)^2 \leq \rs(a \compose b, a \compose b) \cdot \rs(c,c)$.
\item \textbf{Enrichment}: $\rs(a \compose b, a \compose b) \geq \rs(a, a)$.
\end{enumerate}

\section*{Key Definitions}

The \emph{emergence term} is a derived quantity:
\[
\emerge(a, b, c) := \rs(a \compose b, c) - \rs(a, c) - \rs(b, c).
\]
The \emph{weight} of an idea is $w(a) := \rs(a, a)$.
Two ideas are \emph{orthogonal} if $\rs(a, b) = 0$ and $\rs(b, a) = 0$.
An agent is \emph{left-linear} if $\emerge(a, b, c) = 0$ for all $b, c$,
meaning their contributions never produce emergence.

\section*{Key Results from IDT}

The following results, proved in the predecessor volume, will be used freely:

\begin{itemize}
\item \textbf{Cocycle condition}: $\emerge(a \compose b, c, d) = \emerge(a, b \compose c, d) - \emerge(a, b, d) + \emerge(b, c, d)$.
\item \textbf{Decomposition}: $\rs(a \compose b, c) = \rs(a, c) + \rs(b, c) + \emerge(a, b, c)$.
\item \textbf{Void emergence}: $\emerge(\void, a, b) = \emerge(a, \void, b) = \emerge(a, b, \void) = 0$.
\item \textbf{Weight ratchet}: $w(a \compose b) \geq w(a)$ for all $a, b$.
\item \textbf{Non-degeneracy}: $w(a) > 0$ for all $a \neq \void$.
\end{itemize}

These are not assumptions---they are theorems derived from the eight axioms.
The proofs appear in \emph{Idea Diffusion Theory} and in the Lean 4 files.

\vspace{1em}
\noindent\textit{With these tools in hand, we turn to the strategic question:
How should you choose what to say?}

%==========================================================================
\part{The Structure of Meaning Games}
%==========================================================================

\input{tex_v2_chapters/mg_ch1_7}

%==========================================================================
\part{Dynamic and Evolutionary Games}
%==========================================================================

\input{tex_v2_chapters/mg_ch8_14}

%==========================================================================
\part{Advanced Topics and Open Problems}
%==========================================================================

\input{tex_v2_chapters/mg_ch15_21}

%==========================================================================
\part{Deep Structures and Grand Unification}
%==========================================================================

\input{tex_v2_chapters/mg_deep}

%==========================================================================
% END MATTER
%==========================================================================

\chapter*{Appendix: Complete Lean 4 Formalization}

All game-theoretic results in this book are verified in Lean 4.
The relevant files include:

\begin{itemize}
\item \texttt{IDT\_v3\_SignalGames.lean} --- 208 theorems. Nash equilibrium,
babbling equilibria, costly signaling, revelation, strategic silence,
signaling chains, grand unification.

\item \texttt{IDT\_v3\_Wittgenstein.lean} --- 183 theorems. Language games,
rule-following, private language, family resemblance, hinge propositions,
Tractatus vs.\ Investigations, showing vs.\ saying, philosophical therapy.

\item \texttt{IDT\_v3\_DeepGames.lean} --- 72 theorems. Transparency,
order sensitivity, attribution impossibility, weight ratchet, coalitions,
fundamental synthesis.

\item \texttt{IDT\_v3\_Interpretation.lean} --- 230 theorems. Gadamer,
Ricoeur, Schleiermacher, Benjamin, Heidegger, Derrida, hermeneutic circle.

\item \texttt{IDT\_v3\_Semiotics.lean} --- 158 theorems. Eco, Barthes,
Derrida, Jakobson, Greimas.

\item \texttt{IDT\_v3\_PowerStructure.lean} --- 152 theorems. Bourdieu,
Gramsci, Foucault, Althusser, hegemony, ideology, propaganda, resistance.

\item \texttt{IDT\_v3\_Cooperation.lean} --- 26 theorems. Coalition value,
cooperation surplus, Shapley values.

\item \texttt{IDT\_v3\_Evolution.lean} --- 24 theorems. Fitness, invasion,
Red Queen, memetic drift.

\item \texttt{IDT\_v3\_Poetics.lean} --- 103 theorems. Condensation,
defamiliarization, meter, rhyme.

\item \texttt{IDT\_v3\_Emergence.lean} --- 199 theorems. Cocycle algebra,
curvature, stability, differential calculus.

\item \texttt{IDT\_v3\_Networks.lean} --- 120 theorems. Network topology,
diffusion dynamics, cascades.
\end{itemize}

\noindent\textbf{Total: 1,827 theorems across 18 files, all compiling with zero
\texttt{sorry} axioms.}

\chapter*{Bibliography}

\begin{enumerate}[label={[\arabic*]}]
\item Aristotle, \emph{Rhetoric}.
\item Austin, J.L., \emph{How to Do Things with Words}, Oxford, 1962.
\item Bakhtin, M.M., \emph{The Dialogic Imagination}, University of Texas Press, 1981.
\item Barthes, R., \emph{S/Z}, Hill and Wang, 1974.
\item Barthes, R., ``The Death of the Author,'' 1967.
\item Bourdieu, P., \emph{Distinction: A Social Critique of the Judgement of Taste}, Harvard, 1984.
\item Crawford, V.\ and Sobel, J., ``Strategic Information Transmission,''
\emph{Econometrica} 50(6), 1982.
\item Dawkins, R., \emph{The Selfish Gene}, Oxford, 1976.
\item Derrida, J., \emph{Of Grammatology}, Johns Hopkins, 1976.
\item Eco, U., \emph{The Open Work}, Harvard, 1989.
\item Foucault, M., \emph{The Order of Things}, Vintage, 1970.
\item Freire, P., \emph{Pedagogy of the Oppressed}, Continuum, 1970.
\item Gadamer, H.-G., \emph{Truth and Method}, Continuum, 1975.
\item Gramsci, A., \emph{Selections from the Prison Notebooks}, International, 1971.
\item Greimas, A.J., \emph{Structural Semantics}, Nebraska, 1983.
\item Grice, H.P., ``Logic and Conversation,'' in \emph{Syntax and Semantics 3}, 1975.
\item Harsanyi, J., ``Games with Incomplete Information,'' \emph{Management Science} 14, 1967--68.
\item Hegel, G.W.F., \emph{Phenomenology of Spirit}, Oxford, 1807/1977.
\item Heidegger, M., \emph{Being and Time}, Harper \& Row, 1962.
\item Husserl, E., \emph{Logical Investigations}, Routledge, 1900/2001.
\item Jakobson, R., ``Closing Statement: Linguistics and Poetics,'' in \emph{Style in Language}, 1960.
\item Lewis, D., \emph{Convention}, Harvard, 1969.
\item McKee, R., \emph{Story}, ReganBooks, 1997.
\item Merleau-Ponty, M., \emph{Phenomenology of Perception}, Routledge, 1945/2012.
\item Myerson, R., \emph{Game Theory: Analysis of Conflict}, Harvard, 1991.
\item Nash, J., ``Non-Cooperative Games,'' \emph{Annals of Mathematics} 54(2), 1951.
\item Osborne, M.\ and Rubinstein, A., \emph{A Course in Game Theory}, MIT Press, 1994.
\item Peirce, C.S., \emph{Collected Papers}, Harvard, 1931--1958.
\item Ricoeur, P., \emph{Interpretation Theory}, TCU Press, 1976.
\item de Saussure, F., \emph{Course in General Linguistics}, Open Court, 1916/1986.
\item Schelling, T., \emph{The Strategy of Conflict}, Harvard, 1960.
\item Schleiermacher, F., \emph{Hermeneutics and Criticism}, Cambridge, 1998.
\item Shannon, C.E., ``A Mathematical Theory of Communication,'' \emph{Bell System Technical Journal}, 1948.
\item Shklovsky, V., ``Art as Technique,'' 1917.
\item Spence, M., ``Job Market Signaling,'' \emph{QJE} 87(3), 1973.
\item von Neumann, J.\ and Morgenstern, O., \emph{Theory of Games and Economic Behavior}, Princeton, 1944.
\item Wittgenstein, L., \emph{Philosophical Investigations}, Blackwell, 1953.
\item Wittgenstein, L., \emph{Tractatus Logico-Philosophicus}, Routledge, 1921/1922.
\end{enumerate}

\end{document}
