\documentclass[11pt,oneside]{book}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usepackage{epigraph}
\usepackage{tocloft}
\usepackage{mdframed}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{axiom_def}[theorem]{Axiom}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{interpretation}[theorem]{Interpretation}

\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,noncomputable,class,instance,where,
    variable,open,namespace,section,end,import,structure,inductive,
    match,with,fun,let,in,if,then,else,do,return,have,show,by,
    exact,rfl,simp,linarith,nlinarith,ring,intro,apply,rw,unfold,
    calc,sorry,Prop,Type,Sort,axiom,extends,deriving},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]",
}

\lstset{
  language=Lean4,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{red},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  xleftmargin=1em,
  xrightmargin=1em,
}

\newcommand{\rs}{\mathrm{rs}}
\newcommand{\emerge}{\kappa}
\newcommand{\compose}{\circ}
\newcommand{\void}{\varepsilon}
\newcommand{\IS}{\mathcal{I}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}

% Surprise callout box
\newmdenv[
  linecolor=red!70!black,
  linewidth=1.5pt,
  backgroundcolor=red!5,
  roundcorner=5pt,
  innertopmargin=8pt,
  innerbottommargin=8pt,
  skipabove=12pt,
  skipbelow=12pt
]{surprisebox}

% Philosophical reflection box
\newmdenv[
  linecolor=blue!60!black,
  linewidth=1pt,
  backgroundcolor=blue!3,
  roundcorner=5pt,
  innertopmargin=8pt,
  innerbottommargin=8pt,
  skipabove=12pt,
  skipbelow=12pt
]{reflectionbox}

\title{\Huge\bfseries The Math of Ideas\\[0.3em]
\LARGE Volume III-B: Paradoxes of Mind and Meaning\\[0.5em]
\Large A Sequel to Volume III: Philosophy}
\author{A Formal Treatment with Machine-Verified Proofs\\[0.3em]
\normalsize\textit{Part of the series ``The Math of Ideas''}\\[0.2em]
\normalsize\textit{56 theorems, 0 sorries, all verified in Lean~4}}
\date{}

\begin{document}

\maketitle

\frontmatter

\chapter*{Preface}

\epigraph{The limits of my language mean the limits of my world.}{Ludwig Wittgenstein, \textit{Tractatus} 5.6}

This volume is a sequel to Volume~III of \emph{The Math of Ideas}, which brought the
axioms of the ideatic space into dialogue with Wittgenstein, Husserl, Hegel, and Gadamer.
Where that volume established the basic correspondences---showing how emergence captures
the non-additive character of understanding, how resonance formalizes meaning-in-context,
and how composition models the meeting of ideas---this sequel pushes deeper.

The central question of this volume is: \emph{What are the paradoxes of mind and meaning,
and what does it look like when mathematics takes them seriously?}

We prove 56 theorems in Lean~4, organized into 15 chapters.  Every proof compiles.
There are no \texttt{sorry} placeholders, no axioms beyond the eight that define
the ideatic space $(\IS, \compose, \void, \rs)$.  Each theorem says something that,
on first encounter, seems counterintuitive---even paradoxical.  But the axioms force
these conclusions.  The mathematics is not a metaphor; it \emph{is} the argument.

\medskip

\noindent\textbf{What you will find here:}

\begin{itemize}[leftmargin=2em]
\item \textbf{Chapters 1--3:} The private language argument made quantitative;
  bounds on self-knowledge; the devastating theorem that self-describing ideas
  have \emph{negative} semantic charge.

\item \textbf{Chapters 4--6:} The hermeneutic circle reconceived as a strictly
  monotone spiral; why phenomenological reduction changes the observer;
  the frame problem as infinite background weight.

\item \textbf{Chapters 7--9:} Dialectical history as irreversible accumulation;
  a formal impossibility proof for philosophical zombies;
  aspect perception and gestalt switches bounded by emergence.

\item \textbf{Chapters 10--12:} The rule-following paradox as inevitable emergence;
  translation loss as a mathematical certainty;
  the beetle in the box made quantitative.

\item \textbf{Chapters 13--15:} Intersubjectivity and the necessity of asymmetry;
  the infinite regress of self-interpretation;
  a grand synthesis proving that mind \emph{is} emergence.
\end{itemize}

\medskip

\noindent A word about method.  Each chapter presents its theorems in natural language,
gives the Lean~4 code, and then offers an extended philosophical commentary explaining
\emph{why} the result is surprising.  The ``surprise boxes'' scattered through the text
highlight the moments where the mathematics says something that no philosopher---not
Wittgenstein, not Heidegger, not Husserl---could have anticipated without the formal
framework.

The proofs are short.  Many are only two or three lines of Lean.  This is not a weakness;
it is the point.  The axioms are \emph{so tightly constrained} that deep philosophical
truths follow from elementary algebraic manipulations.  The surprise is not in the
complexity of the proof but in the \emph{content} of the conclusion.

\vfill

\noindent\textit{All proofs verified in Lean~4 with Mathlib. Source file:
\texttt{FormalAnthropology/IDT\_v3b\_DeepPhilosophy.lean}.}

\tableofcontents

\mainmatter

\input{tex_v2_chapters/vol3b_ch1_5}
\input{tex_v2_chapters/vol3b_ch6_10}
\input{tex_v2_chapters/vol3b_ch11_15}

\backmatter

\chapter*{Appendix: The Eight Axioms}

For reference, the complete axiom system of the ideatic space
$(\IS, \compose, \void, \rs)$:

\begin{enumerate}[label=\textbf{A\arabic*.}]
\item \textbf{Composition.} There exists a binary operation $\compose : \IS \times \IS \to \IS$.

\item \textbf{Associativity.} For all $a, b, c \in \IS$:
  $(a \compose b) \compose c = a \compose (b \compose c)$.

\item \textbf{Identity.} There exists an element $\void \in \IS$ such that
  $a \compose \void = a = \void \compose a$ for all $a$.

\item \textbf{Resonance.} There exists a function $\rs : \IS \times \IS \to \R$.

\item \textbf{Silence contributes nothing.}
  $\rs(a, \void) = 0$ and $\rs(\void, a) = 0$ for all $a$.

\item \textbf{Self-resonance is non-negative.}
  $\rs(a, a) \geq 0$ for all $a$.

\item \textbf{Non-degeneracy.} $\rs(a, a) = 0$ if and only if $a = \void$.

\item \textbf{Composition enriches (Cauchy--Schwarz bound).}
  $\rs(a \compose b, a \compose b) \geq \rs(a, a)$ and
  $(\emerge(a,b,c))^2 \leq \rs(a \compose b, a \compose b) \cdot \rs(c, c)$,
  where $\emerge(a,b,c) = \rs(a \compose b, c) - \rs(a,c) - \rs(b,c)$.
\end{enumerate}

\bigskip

\noindent These eight axioms---and nothing more---generate every theorem in this volume.

\chapter*{Appendix: Complete Lean~4 Source}

The complete machine-verified source code for all 56 theorems in this volume is
available in the file:

\begin{center}
\texttt{FormalAnthropology/IDT\_v3b\_DeepPhilosophy.lean}
\end{center}

\noindent The file imports only \texttt{IDT\_v3\_Foundations}, which provides the
\texttt{IdeaticSpace3} type class and derived lemmas.  It compiles with
\texttt{lake build} using Lean~4 and Mathlib, producing zero errors and zero warnings.

\noindent Key statistics:
\begin{itemize}
\item 982 lines of Lean~4 code
\item 56 theorems and lemmas
\item 0 \texttt{sorry} placeholders
\item 15 sections corresponding to the 15 chapters of this volume
\end{itemize}

\end{document}
