\documentclass[11pt,oneside]{book}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usepackage{epigraph}
\usepackage{tocloft}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{axiom_def}[theorem]{Axiom}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{interpretation}[theorem]{Interpretation}

\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,noncomputable,class,instance,where,
    variable,open,namespace,section,end,import,structure,inductive,
    match,with,fun,let,in,if,then,else,do,return,have,show,by,
    exact,rfl,simp,linarith,nlinarith,ring,intro,apply,rw,unfold,
    calc,sorry,Prop,Type,Sort,axiom,extends,deriving},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]",
}

\lstset{
  language=Lean4,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{red},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  xleftmargin=1em,
  xrightmargin=1em,
}

\newcommand{\rs}{\mathrm{rs}}
\newcommand{\emerge}{\kappa}
\newcommand{\compose}{\circ}
\newcommand{\void}{\varepsilon}
\newcommand{\IS}{\mathcal{I}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}

\title{\Huge\bfseries The Math of Ideas\\[0.3em]
\LARGE Volume III: Philosophy\\[0.5em]
\Large Wittgenstein, Hermeneutics, and the Formal Structure of Understanding}
\author{A Formal Treatment with Machine-Verified Proofs\\[0.3em]
\normalsize\textit{Part of the series ``The Math of Ideas''}}
\date{}

\begin{document}

\maketitle

\frontmatter

\chapter*{Preface}

This is the third volume in the series \emph{The Math of Ideas}. Where Volume~I
established the foundations---the eight axioms of the ideatic space
$(\IS, \compose, \void, \rs)$ and the emergence function
$\emerge(a,b,c) = \rs(a \compose b, c) - \rs(a,c) - \rs(b,c)$---this volume
brings those axioms into direct dialogue with the central texts of
philosophy of meaning.

The philosophical tradition has produced some of humanity's deepest insights
about meaning, understanding, and interpretation. Wittgenstein showed that
meaning is use. Gadamer revealed that understanding is always situated.
Heidegger uncovered the structures of being-in-the-world. Derrida demonstrated
that meaning is always deferred. Hegel traced the dialectical development of
ideas through history. Each of these insights has been expressed in prose
of extraordinary subtlety and power. None has been \emph{proved}.

This volume proves them. Or rather: it identifies the precise formal content of
each philosophical claim, states it as a theorem within the IDT framework,
and provides a machine-verified proof. In doing so, it discovers that some
claims are theorems (they follow from the axioms), some are definitions
(they name structures already present in the formalism), and some are
\emph{false}---they require assumptions beyond what the axioms provide.
The last category is perhaps the most valuable: it tells us exactly where
philosophical intuition outruns mathematical structure, and what additional
axioms would be needed to recover the philosopher's original insight.

Every theorem in this volume has been formally verified in Lean~4 with zero
\texttt{sorry} axioms. The Lean source files are:
\begin{itemize}
\item \texttt{IDT\_v3\_Wittgenstein.lean} --- 370 theorems on language games,
rule-following, private language, family resemblance, hinge propositions,
aspect-seeing, meaning blindness, and the Tractatus/Investigations divide.

\item \texttt{IDT\_v3\_Interpretation.lean} --- 478 theorems on fusion of
horizons, iterated reading, the hermeneutic circle, prejudice, interpretive
depth, communication asymmetry, tradition, and intertextuality.

\item \texttt{IDT\_v3\_Phenomenology.lean} --- 197 theorems on intentionality,
horizons, givenness, the reduction, temporality, intersubjectivity,
embodiment, the lifeworld, Heidegger's being-in-the-world, Merleau-Ponty's
flesh, and Levinas's face.

\item \texttt{IDT\_v3\_Dialectics.lean} --- 225 theorems on thesis/antithesis/synthesis,
Aufhebung decomposition, iterated dialectics, Adorno's non-identical,
master-slave, dialectic of enlightenment, and dialectical sequences.
\end{itemize}

The reader is assumed to be familiar with Volume~I, whose results are
cited throughout as ``Theorem~I.\emph{X.Y}.'' No prior knowledge of
formal logic or proof assistants is required, though occasional Lean~4
code snippets are included for those who wish to verify the proofs
directly.

\vspace{1em}
\noindent\textit{A note on philosophical engagement.} This book is not a
\emph{summary} of Wittgenstein, Gadamer, Heidegger, Derrida, Hegel, or
any other philosopher. It is a \emph{reply}. The formalism agrees with
these thinkers on many points, disagrees on others, and in several cases
reveals structural insights that the philosophers themselves could not have
seen without the mathematics. We treat these thinkers as interlocutors,
not as authorities.

\tableofcontents

\mainmatter

%==========================================================================
\part{Wittgenstein}
%==========================================================================

\input{tex_v2_chapters/vol3_ch1_5}

%==========================================================================
\part{Hermeneutics and Phenomenology}
%==========================================================================

\input{tex_v2_chapters/vol3_ch6_10}

%==========================================================================
\part{Dialectics and Synthesis}
%==========================================================================

\input{tex_v2_chapters/vol3_ch11_15}

%==========================================================================
% END MATTER
%==========================================================================

\backmatter

\chapter*{Appendix: Index of Lean Theorems by Philosopher}

The following index maps each philosopher discussed in this volume to the
principal Lean~4 theorems that formalize their ideas. File references use
the prefix \texttt{IDT\_v3\_} throughout.

\begin{description}[style=nextline]
\item[Wittgenstein]
  \texttt{Wittgenstein.lean}: \texttt{gameMove\_void\_player} (W1),
  \texttt{gameScore\_decompose} (W5),
  \texttt{formOfLife\_absorb} (W15),
  \texttt{familyRelated\_self} (W26),
  \texttt{family\_resemblance\_non\_transitive} (W113),
  \texttt{private\_language\_argument} (W51),
  \texttt{beetle\_drops\_out} (W63),
  \texttt{publiclyInvisible\_is\_void} (W59),
  \texttt{rule\_paradox\_same\_output} (W42),
  \texttt{community\_decides} (W103),
  \texttt{void\_is\_hinge} (W133),
  \texttt{duck\_rabbit} (W72),
  \texttt{meaning\_eq\_said\_plus\_shown} (W156),
  \texttt{tractarian\_compositional} (W152),
  \texttt{investigations\_beyond\_tractatus} (W153).

\item[Gadamer]
  \texttt{Interpretation.lean}: \texttt{fusionOfHorizons} (def),
  \texttt{fusion\_enriches\_reader},
  \texttt{fusion\_assoc},
  \texttt{hermeneutic\_circle\_resolves},
  \texttt{hermeneutic\_total\_emergence},
  \texttt{void\_prejudice\_no\_emergence},
  \texttt{compositional\_prejudice},
  \texttt{readN\_enriches},
  \texttt{perfect\_communication\_iff}.

\item[Husserl]
  \texttt{Phenomenology.lean}: \texttt{intentionalAct\_eq},
  \texttt{intentionality\_enriches},
  \texttt{noematic\_decomposition},
  \texttt{noetic\_decomposition},
  \texttt{reduction\_decomposition},
  \texttt{passiveSynthesis\_symmetric},
  \texttt{resonance\_passive\_active},
  \texttt{layered\_constitution},
  \texttt{sharesEidos\_iff\_emergence}.

\item[Heidegger]
  \texttt{Phenomenology.lean}: \texttt{dasein\_empty\_world},
  \texttt{dasein\_no\_projection},
  \texttt{beingTowardDeath\_decomposition},
  \texttt{readyToHand\_linear},
  \texttt{clearing\_cocycle},
  \texttt{thrownness\_nonneg},
  \texttt{ecstatic\_temporality}.

\item[Hegel]
  \texttt{Dialectics.lean}: \texttt{synthesis\_enriches\_thesis},
  \texttt{aufhebung\_decomposition},
  \texttt{dialectical\_cocycle},
  \texttt{iteratedSynthesis\_enriches},
  \texttt{dialecticalSequence\_enriches},
  \texttt{negation\_of\_negation\_enriches}.

\item[Adorno]
  \texttt{Dialectics.lean}: \texttt{nonIdentical},
  \texttt{resistance\_eq},
  \texttt{adorno\_balance},
  \texttt{negation\_residue\_nonneg}.

\item[Derrida]
  \texttt{Semiotics.lean}: \texttt{differance},
  \texttt{derrida\_trace},
  \texttt{supplement\_void}.
  \texttt{Dialectics.lean}: \texttt{reificationDeficit\_eq\_aufhebung}.

\item[Levinas]
  \texttt{Phenomenology.lean}: \texttt{face\_excess\_bounded},
  \texttt{totality\_plus\_infinity},
  \texttt{saying\_and\_said},
  \texttt{ethicalSurplus\_lower\_bound},
  \texttt{substitution\_antisymmetric}.

\item[Merleau-Ponty]
  \texttt{Phenomenology.lean}: \texttt{chiasm\_antisymmetric},
  \texttt{reversible\_symmetric},
  \texttt{motorIntentionality\_decomposition},
  \texttt{intercorporeality\_cocycle}.
\end{description}

\end{document}
