\documentclass[11pt,oneside]{book}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usepackage{epigraph}
\usepackage{tocloft}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{axiom_def}[theorem]{Axiom}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{interpretation}[theorem]{Interpretation}

\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,noncomputable,class,instance,where,
    variable,open,namespace,section,end,import,structure,inductive,
    match,with,fun,let,in,if,then,else,do,return,have,show,by,
    exact,rfl,simp,linarith,nlinarith,ring,intro,apply,rw,unfold,
    calc,sorry,Prop,Type,Sort,axiom,extends,deriving},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]",
}

\lstset{
  language=Lean4,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{red},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  xleftmargin=1em,
  xrightmargin=1em,
  literate={→}{{$\to$}}1 {←}{{$\leftarrow$}}1 {↦}{{$\mapsto$}}1
           {∀}{{$\forall$}}1 {∃}{{$\exists$}}1 {λ}{{$\lambda$}}1
           {≤}{{$\leq$}}1 {≥}{{$\geq$}}1 {≠}{{$\neq$}}1
           {∧}{{$\wedge$}}1 {∨}{{$\vee$}}1 {¬}{{$\neg$}}1
           {∈}{{$\in$}}1 {∉}{{$\notin$}}1 {⊆}{{$\subseteq$}}1
           {ℕ}{{$\mathbb{N}$}}1 {ℤ}{{$\mathbb{Z}$}}1 {ℝ}{{$\mathbb{R}$}}1
           {∘}{{$\circ$}}1 {×}{{$\times$}}1 {⟨}{{$\langle$}}1 {⟩}{{$\rangle$}}1
           {↔}{{$\leftrightarrow$}}1 {·}{{$\cdot$}}1 {ε}{{$\varepsilon$}}1
           {∩}{{$\cap$}}1 {₀}{{$_0$}}1,
}

\newcommand{\rs}{\mathrm{rs}}
\newcommand{\emerge}{\kappa}
\newcommand{\compose}{\circ}
\newcommand{\void}{\varepsilon}
\newcommand{\IS}{\mathcal{I}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}

\title{\Huge\bfseries The Math of Ideas\\[0.3em]
\LARGE Volume IV: Signs and Culture\\[0.5em]
\Large Semiotics, Poetics, Narrative, and Translation}
\author{A Formal Treatment with Machine-Verified Proofs\\[0.3em]
\normalsize\textit{Part of the series ``The Math of Ideas''}}
\date{}

\begin{document}
\maketitle

% ================================================================
% PREFACE
% ================================================================
\frontmatter

\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

\epigraph{The limits of my language mean the limits of my world.}{Ludwig Wittgenstein, \textit{Tractatus Logico-Philosophicus}}

This is the fourth volume of \textit{The Math of Ideas}, a series that develops
a rigorous mathematical framework for the phenomena traditionally studied by the
humanities --- meaning, culture, narrative, aesthetics, translation, and social
life. Volume~I established the foundations: the eight axioms of an
\textit{IdeaticSpace} that capture the algebraic structure of ideas, their
composition, and the emergence of new meaning from combination. Volume~IV takes
those same eight axioms and develops from them a \textit{formal semiotics} ---
the mathematics of signs, codes, literary form, narrative structure, translation,
and aesthetic experience.

The ambition is immodest: to show that the central insights of Saussure, Peirce,
Barthes, Eco, Lotman, Shklovsky, Jakobson, Propp, Bakhtin, Genette, Ricoeur,
Benjamin, Venuti, Kant, and Adorno can be captured as theorems within a single
deductive system, and that these theorems can be \textit{machine-verified} in
the Lean~4 proof assistant. This is not an act of reductionism --- reducing the
richness of literature and culture to dry formalism --- but an act of
\textit{translation}: rendering the profound intuitions of these thinkers into a
language where their logical relationships become visible, their hidden
assumptions become explicit, and their consequences can be rigorously derived.

\subsection*{What This Volume Covers}

The volume is organized into three parts:

\begin{itemize}[leftmargin=2em]
\item \textbf{Part~I: Formal Semiotics} (Chapters~1--5) develops the
  mathematics of signs. We formalize Saussure's arbitrariness of the sign,
  Peirce's triadic semiosis, Barthes's theory of myth, Eco's open and closed
  works, and Lotman's semiosphere --- all as consequences of the eight axioms.

\item \textbf{Part~II: Poetics and Narrative} (Chapters~6--10) turns to literary
  form. Shklovsky's defamiliarization, Jakobson's poetic function, the full
  apparatus of rhetorical figures (metaphor, metonymy, irony, hyperbole),
  narrative structure from Propp through Ricoeur, Bakhtin's heteroglossia, and
  the formal constraints of verse --- all receive axiomatic treatment.

\item \textbf{Part~III: Translation and Aesthetics} (Chapters~11--15) addresses
  the movement of meaning across languages and the nature of aesthetic
  experience. Benjamin's theory of translation, the problem of
  untranslatability, Kant's aesthetics of the beautiful and sublime, Adorno's
  dialectical aesthetics, and a synthetic mathematical aesthetics are developed
  as theorems.
\end{itemize}

\subsection*{The Lean Proofs}

Every theorem in this volume has been machine-verified in Lean~4. The proofs are
contained in five files:

\begin{center}
\begin{tabular}{llr}
\textbf{File} & \textbf{Domain} & \textbf{Theorems} \\
\hline
\texttt{IDT\_v3\_Semiotics.lean} & Signs, codes, ideology & 327 \\
\texttt{IDT\_v3\_Poetics.lean} & Verse, rhetoric, figuration & 251 \\
\texttt{IDT\_v3\_Narrative.lean} & Story, plot, narrative theory & 240 \\
\texttt{IDT\_v3\_Translation.lean} & Translation, fidelity, equivalence & 301 \\
\texttt{IDT\_v3\_Aesthetics.lean} & Beauty, sublimity, taste & 207 \\
\hline
& \textbf{Total} & \textbf{1,326}
\end{tabular}
\end{center}

All five files import \texttt{IDT\_v3\_Foundations.lean}, which provides the
eight axioms. No additional axioms are introduced; every result in this volume
is a consequence of the foundational eight.

\subsection*{How to Read This Volume}

The reader who has worked through Volume~I will find the mathematical language
familiar. We use $\IS$ for an IdeaticSpace, $\compose$ for composition,
$\void$ for the null utterance, $\rs(a, b)$ for resonance, and
$\emerge(a, b, c) = \rs(a \compose b, c) - \rs(a, c) - \rs(b, c)$ for
emergence. The eight axioms are recalled below for convenience.

The reader approaching from the humanities --- a literary theorist, a
semiotician, a translator --- should not be intimidated by the formalism. Each
theorem is followed by an \textit{Interpretation} that explains its significance
in plain language and connects it to the tradition. The Lean code listings are
provided for verification but can be skipped without loss of understanding.

The reader approaching from mathematics or computer science will find a rich
source of examples of how abstract algebra (monoids, cocycles, equivalence
relations, group actions) arises naturally in the analysis of cultural
phenomena.

\subsection*{The Eight Axioms}

For reference, the eight axioms of an IdeaticSpace, as established in Volume~I:

\begin{axiom_def}[A1: Associativity]
For all ideas $a, b, c \in \IS$:
$\quad (a \compose b) \compose c = a \compose (b \compose c)$
\end{axiom_def}

\begin{axiom_def}[A2: Left Identity]
For all $a \in \IS$: $\quad \void \compose a = a$
\end{axiom_def}

\begin{axiom_def}[A3: Right Identity]
For all $a \in \IS$: $\quad a \compose \void = a$
\end{axiom_def}

\begin{axiom_def}[R1: Void Target Annihilation]
For all $a \in \IS$: $\quad \rs(a, \void) = 0$
\end{axiom_def}

\begin{axiom_def}[R2: Void Source Annihilation]
For all $a \in \IS$: $\quad \rs(\void, a) = 0$
\end{axiom_def}

\begin{axiom_def}[E1: Self-Resonance Non-Negativity]
For all $a \in \IS$: $\quad \rs(a, a) \geq 0$
\end{axiom_def}

\begin{axiom_def}[E2: Nondegeneracy]
For all $a \in \IS$: $\quad \rs(a, a) = 0 \implies a = \void$
\end{axiom_def}

\begin{axiom_def}[E3: Emergence Boundedness]
For all $a, b, c \in \IS$:
$\quad \emerge(a, b, c)^2 \leq \rs(a \compose b, a \compose b) \cdot \rs(c, c)$
\end{axiom_def}

\begin{axiom_def}[E4: Composition Enriches]
For all $a, b \in \IS$:
$\quad \rs(a \compose b, a \compose b) \geq \rs(a, a)$
\end{axiom_def}

\vspace{1em}

These eight axioms --- three algebraic, two annihilation, four analytic ---
generate the entire theory. Every theorem in this volume, from the arbitrariness
of the sign to the impossibility of perfect translation to the formal structure
of beauty, is a logical consequence of these eight statements.

\vspace{2em}
\hfill \textit{--- The Author, 2026}

\tableofcontents

% ================================================================
% MAIN MATTER
% ================================================================
\mainmatter

% ----------------------------------------------------------------
% PART I: FORMAL SEMIOTICS
% ----------------------------------------------------------------
\part{Formal Semiotics}
\label{part:semiotics}

\input{tex_v2_chapters/vol4_ch1_5}

% ----------------------------------------------------------------
% PART II: POETICS AND NARRATIVE
% ----------------------------------------------------------------
\part{Poetics and Narrative}
\label{part:poetics}

\input{tex_v2_chapters/vol4_ch6_10}

% ----------------------------------------------------------------
% PART III: TRANSLATION AND AESTHETICS
% ----------------------------------------------------------------
\part{Translation and Aesthetics}
\label{part:translation}

\input{tex_v2_chapters/vol4_ch11_15}

% ================================================================
% BACK MATTER
% ================================================================
\backmatter

\chapter*{Afterword: What Formalization Reveals}
\addcontentsline{toc}{chapter}{Afterword}

\epigraph{Mathematics is the art of giving the same name to different things.}{Henri Poincar\'{e}}

We have traveled a long arc in this volume --- from Saussure's insight that
meaning is differential, through the triadic semiosis of Peirce, the mythologies
of Barthes, the narrative structures of Propp and Ricoeur, the agonies of
translation, and finally to the formal structure of beauty itself. Along the
way, a single mathematical concept has served as our guiding thread:
\textit{emergence}.

The emergence function $\emerge(a, b, c) = \rs(a \compose b, c) - \rs(a, c) -
\rs(b, c)$ measures the irreducible surplus that arises when ideas are combined.
It is this surplus that makes a sign more than a label (Chapter~1), an index
more than a correlation (Chapter~2), a myth more than a statement (Chapter~3),
a poem more than a sequence of words (Chapters~6--10), a translation more than
a substitution cipher (Chapter~11), and a work of art more than an artifact
(Chapters~13--15).

What has formalization revealed that was not already apparent?

\textbf{First}, that the seemingly disparate insights of twentieth-century
semiotics, literary theory, translation studies, and aesthetics are
\textit{logically connected}. Saussure's arbitrariness and Peirce's iconicity
are not competing theories but complementary consequences of the same axioms
(Chapters~1--2). Barthes's myth and Eco's open work are both about emergence,
differing only in whether it is suppressed or celebrated (Chapters~3--4).
Benjamin's ``pure language'' and Kant's ``purposiveness without purpose'' are
both limit concepts in the same emergence framework (Chapters~11, 13).

\textbf{Second}, that certain impossibility results are genuine theorems, not
mere pessimism. The impossibility of perfectly faithful non-trivial translation
(Theorem~11.6), the exclusivity of iconicity and symbolic reference
(Theorem~2.5), the zero originality of kitsch (Theorem~14.3) --- these are not
opinions but mathematical necessities, given the axioms.

\textbf{Third}, that the cocycle condition --- the associativity constraint on
emergence that falls out of Axiom~A1 --- is the hidden structural law governing
narrative (the three-act cocycle, the monomyth cocycle, Ricoeur's mimesis
cocycle), translation (the chain distortion decomposition), and aesthetic
experience (the dialectic cocycle, the habituation cocycle). Wherever meaning is
built up through sequential composition, the cocycle constrains what is possible.

\textbf{Fourth}, that the vocabulary of formal mathematics --- equivalence
relations, group actions, triangle inequalities, fixed points, conservation
laws --- maps onto the vocabulary of cultural theory with startling precision.
Taste is an equivalence relation. Translation is a group action. Aesthetic
proximity satisfies a triangle inequality. The untranslatable is a fixed point
of the identity. These are not metaphors; they are theorems.

We do not claim that this formalization exhausts the meaning of the works we
have engaged. Saussure's \textit{Cours} is more than Theorem~1.1; the
\textit{Recherche} is more than a cocycle. But we do claim that the formal
skeleton we have exposed is genuinely \textit{there} --- that it is part of
what these works mean, and that seeing it clearly enriches rather than diminishes
our understanding.

The 1,326 machine-verified theorems in this volume constitute, we believe, the
largest body of formally proved results in the humanities. They are offered not
as the last word but as the first word in a new kind of conversation between
mathematics and culture.

\vspace{2em}

\chapter*{Notation Index}
\addcontentsline{toc}{chapter}{Notation Index}

\begin{tabular}{lll}
\textbf{Symbol} & \textbf{Name} & \textbf{Definition} \\
\hline
$\IS$ & Ideatic space & The space of ideas \\
$\compose$ & Composition & Binary operation on ideas \\
$\void$ & Void & The null utterance / silence \\
$\rs(a, b)$ & Resonance & How $a$ resonates with $b$ \\
$\emerge(a, b, c)$ & Emergence & $\rs(a \compose b, c) - \rs(a, c) - \rs(b, c)$ \\
$w(a)$ & Weight & $\rs(a, a)$, self-resonance \\
$a^{\langle n \rangle}$ & $n$-fold composition & $\mathrm{composeN}(a, n)$ \\
$\phi, \psi$ & Translation maps & Functions $\IS \to \IS$ \\
$\mathrm{fid}(\phi, a, b)$ & Translation fidelity & $\rs(\phi(a), \phi(b)) - \rs(a, b)$ \\
\end{tabular}

\vspace{2em}

\chapter*{Lean File Reference}
\addcontentsline{toc}{chapter}{Lean File Reference}

All proofs are machine-verified in Lean~4 using Mathlib. The source files are
organized as follows:

\begin{description}[leftmargin=3em]
\item[\texttt{IDT\_v3\_Foundations.lean}] The eight axioms (class
  \texttt{IdeaticSpace}) and basic derived notions: emergence, weight,
  composeN, sameIdea, leftLinear, polysemous.

\item[\texttt{IDT\_v3\_Semiotics.lean}] 327~theorems. Saussure (arbitrariness,
  connotation, structural opposition, minimal pairs, paradigmatic/syntagmatic).
  Peirce (icon, index, symbol, sign irreducibility). Barthes (myth, naturalization,
  third-order signs). Eco (open/closed texts). Lotman (semiosphere, frozen/living
  signs, hegemony, ethical encounter). Digital semiotics (emoji, meme, hashtag).
  Biosemiotics (umwelt, functional tone). Grand semiotic dichotomy and master
  cocycle.

\item[\texttt{IDT\_v3\_Poetics.lean}] 251~theorems. Repetition and anaphora.
  Meter (iamb, trochee, spondee, regular meter). Rhyme (perfect, near, strength).
  Compression and condensation. Variation and parallelism. Chiasmus and
  antithesis. Metaphor, metonymy, synecdoche, hyperbole, litotes, oxymoron,
  zeugma. Sonnet structure (Shakespearean, Petrarchan, volta). Enjambment.
  Prosody (rising/falling rhythm, caesura). Stanza and couplet. Tragic-sublime
  link.

\item[\texttt{IDT\_v3\_Narrative.lean}] 240~theorems. Narrative weight and
  emergence. Defamiliarization and dramatic tension. Rhetoric (signal, context,
  pathos). Three-act structure. Climax and closure. Propp's functions. Hero's
  journey and monomyth. Barthes's five codes. Focalization and temporal
  asymmetry. Narrator reliability and dramatic irony. Chekhov's gun. Frame
  narratives. Intertextuality. Ricoeur's mimesis. Masterplot. Iser's reader
  response. Derrida's trace. Kristeva's transposition. Greimas's actantial model.
  Jakobson's functions. Spivak's subaltern. Narrative entropy.

\item[\texttt{IDT\_v3\_Aesthetics.lean}] 207~theorems. Beauty, ugliness, and
  neutrality. The sublime. Taste as equivalence relation. Originality and kitsch.
  Camp. Aesthetic surprise and defamiliarization. Benjamin's aura. Adorno's
  critique and dialectic. Rancière's distribution of the sensible. Bourdieu's
  cultural capital and symbolic violence. Aesthetic experience. Artistic form and
  coherence. Recursive art. Co-creation. Aesthetic resistance. Paradigm shift.
  Aesthetic depth and completeness.
\end{description}

\end{document}
