\documentclass[11pt,oneside]{book}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathrsfs}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{tikz}
\usepackage{epigraph}
\usepackage{tocloft}

% Theorem environments
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[chapter]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{axiom_def}[theorem]{Axiom}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{interpretation}[theorem]{Interpretation}

% Lean code listings
\lstdefinelanguage{Lean4}{
  morekeywords={theorem,lemma,def,noncomputable,class,instance,where,
    variable,open,namespace,section,end,import,structure,inductive,
    match,with,fun,let,in,if,then,else,do,return,have,show,by,
    exact,rfl,simp,linarith,nlinarith,ring,intro,apply,rw,unfold,
    calc,sorry,Prop,Type,Sort,axiom,extends,deriving},
  sensitive=true,
  morecomment=[l]{--},
  morecomment=[s]{/-}{-/},
  morestring=[b]",
}

\lstset{
  language=Lean4,
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{red},
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!5},
  xleftmargin=1em,
  xrightmargin=1em,
}

% Custom commands
\newcommand{\rs}{\mathrm{rs}}
\newcommand{\emerge}{\kappa}
\newcommand{\compose}{\circ}
\newcommand{\void}{\varepsilon}
\newcommand{\IS}{\mathcal{I}}
\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}

\title{\Huge\bfseries The Math of Ideas\\[0.3em]
\LARGE Volume V: Power, Knowledge, and Ethics\\[0.5em]
\Large Critical Theory, Epistemology, and Moral Philosophy in Ideatic Space}
\author{A Formal Treatment with Machine-Verified Proofs\\[0.3em]
\normalsize\textit{Part of the series ``The Math of Ideas''}}
\date{}

\begin{document}

\maketitle

\frontmatter

\chapter*{Preface}

This is the fifth volume in the series \emph{The Math of Ideas}. The first volume
established the foundations: eight axioms governing how ideas compose, resonate,
and give rise to emergence. Volume~II studied games and strategic interaction;
Volume~III examined the philosophy of meaning; Volume~IV formalized semiotics
and cultural transmission. This volume turns to what is arguably the most
consequential domain of all: \emph{power, knowledge, and ethics}---three
dimensions that, we shall argue, are not separate subjects but three perspectives
on a single mathematical phenomenon.

The central thesis of this volume is that \textbf{power, knowledge, and ethics
are three perspectives on how ideas compose with asymmetric resonance}. When
one idea dominates another in self-resonance, we call it \emph{power}. When
ideas compose to form justified beliefs, we call the process \emph{knowledge}.
When we ask how the surplus created by composition---the emergence---should be
distributed, we enter the domain of \emph{ethics}. But mathematically, these
are all instances of the same operations: composition, resonance, and emergence.

This unification is not a rhetorical gesture. It is a proven mathematical fact.
Every theorem in this volume is derived from the same eight axioms established in
Volume~I, and every proof has been verified in Lean~4 with zero \texttt{sorry}
axioms. The Lean formalization comprises:

\begin{itemize}
\item \texttt{IDT\_v3\_PowerStructure.lean} --- 335 theorems on power, hegemony,
ideology, domination, censorship, and institutional structure.
\item \texttt{IDT\_v3\_Ethics.lean} --- 199 theorems on moral weight, virtue,
utility, justice, care, dilemmas, and responsibility.
\item \texttt{IDT\_v3\_Epistemology.lean} --- 118 theorems on belief, knowledge,
justification, paradigms, coherence, and epistemic injustice.
\item \texttt{IDT\_v3\_Diffusion.lean} --- 319 theorems on transmission,
reception, propaganda, echo chambers, polarization, and media.
\end{itemize}

This is not a textbook \emph{about} Gramsci, Foucault, Rawls, or Kuhn. It is a
\emph{reply} to them---a demonstration that their deepest insights can be stated
as theorems and their arguments can be made rigorous. Where they relied on
metaphor, we provide proof. Where they gestured at structure, we exhibit it.
And where they disagreed, the mathematics sometimes reveals that they were
saying the same thing in different notation.

\vspace{1em}

The volume is organized in three parts:

\begin{enumerate}
\item \textbf{Part I: Power and Ideology} (Chapters 1--5) formalizes the
political theory of Gramsci, Foucault, Althusser, Bourdieu, and the theory
of propaganda and censorship.

\item \textbf{Part II: Epistemology} (Chapters 6--10) formalizes the
epistemology of Kuhn, Popper, Quine, and the social epistemologists,
including the theory of epistemic injustice.

\item \textbf{Part III: Ethics} (Chapters 11--15) formalizes the moral
philosophy of Kant, Mill, Aristotle, and Rawls, culminating in our own
theory of the ethics of ideas.
\end{enumerate}

\tableofcontents

\mainmatter

%==========================================================================
\part{Power and Ideology}
%==========================================================================

\input{tex_v2_chapters/vol5_ch1_5}

%==========================================================================
\part{Epistemology}
%==========================================================================

\input{tex_v2_chapters/vol5_ch6_10}

%==========================================================================
\part{Ethics}
%==========================================================================

\input{tex_v2_chapters/vol5_ch11_15}

%==========================================================================
% END MATTER
%==========================================================================

\chapter*{Appendix: Complete Lean 4 Formalization}

All theorems in this book have been formally verified in Lean~4 using the
\texttt{IdeaticSpace} type class defined in \texttt{IDT\_v3\_Foundations.lean}.
The formalization for this volume comprises the following files:

\begin{itemize}
\item \texttt{IDT\_v3\_PowerStructure.lean} --- 335 theorems. Power relations,
hegemony, symbolic capital, ideology, propaganda, resistance, biopower,
censorship, interpellation, structural violence, Arendt's power/violence
distinction, Lukes's three dimensions, Weber's authority types, Butler's
performativity, Habermas's colonization, Mouffe's agonism, Scott's hidden
transcripts, Fanon's decolonization, and institutional analysis.

\item \texttt{IDT\_v3\_Ethics.lean} --- 199 theorems. Moral weight, utility,
universalizability, virtue and character, care, the veil of ignorance,
moral disagreement and synthesis, cooperation surplus, attribution
impossibility, act/rule utilitarianism, deontological residue, moral luck,
self-constitution, contractualism, Parfit's personal identity, capability,
moral dilemmas, trolley problems, moral realism/anti-realism, Kant's
categorical imperative, moral generalism/particularism, incommensurability,
supererogation, Kohlberg's stages, reactive attitudes, collective
responsibility, intergenerational justice, demandingness, is/ought gap,
thick concepts, moral perception, tolerance, authority, communitarianism,
eudaimonia, dirty hands, rights, and the utility monster.

\item \texttt{IDT\_v3\_Epistemology.lean} --- 118 theorems. Belief, strong
belief, doubt, suspension of judgment, doxastic capacity, justification,
JTB knowledge, self-knowledge, epistemic luck, Gettier cases, foundationalism,
coherentism, paradigm shifts, incommensurability, Kuhn loss, falsification,
corroboration, Quinean revision and holism, underdetermination, Bayesian
updating, testimony, trust, credibility deficit, testimonial injustice,
hermeneutical gap, collective epistemology, radical skepticism, a priori/a
posteriori, synthetic a priori, intellectual humility, open-mindedness,
epistemic resilience, and iterated knowledge.

\item \texttt{IDT\_v3\_Diffusion.lean} --- 319 theorems. Reception, transmission
emergence, chain transmission, resonance shift, preservation, population
resonance, agreement, conservative transmission, fidelity, iterated
reception, fixed points, canonical ideas, contagion models, broadcast,
signal degradation/amplification, gatekeeping, super-spreaders, vaccination,
disagreement, polarization, echo chambers, filter bubbles, network
transmission, distortion chains, consensus, viral potency, cascade
dynamics, resistance/absorption, semantic drift, competitive advantage,
path dependence, channel noise, McLuhan's media theory, Debord's spectacle,
Chomsky's five filters, adoption thresholds, influence, group polarization,
affective polarization, diffusion kernels, mixing coefficients, ergodic
dynamics, mutation, and thermodynamic analogies.
\end{itemize}

\end{document}
