% ══════════════════════════════════════════════════════════════════════
% Chapter 1: Hermeneutic Foundations
% ══════════════════════════════════════════════════════════════════════

\chapter{Hermeneutic Foundations: Repertoires, Interpretation, and the Hermeneutic Bound}
\label{ch:hermeneutic}

\epigraph{``The limits of my language mean the limits of my world.''}{Ludwig Wittgenstein, \textit{Tractatus Logico-Philosophicus} (5.6)}

% ── Historical Preamble ──────────────────────────────────────────────

The problem of interpretation---how one mind grasps the meaning
produced by another---has occupied Western thought since at least the
Alexandrian grammarians.  Yet it was Friedrich Schleiermacher who, in his
lectures of 1819--1833, first elevated hermeneutics from a collection of
ad hoc philological techniques to a general theory of
understanding.\footnote{For a comprehensive survey of this trajectory,
see Richard E.\ Palmer, \emph{Hermeneutics: Interpretation Theory in
Schleiermacher, Dilthey, Heidegger, and Gadamer} (Evanston: Northwestern
University Press, 1969).}  Schleiermacher's \emph{hermeneutic circle}---the
thesis that one cannot understand the parts of a text without
understanding the whole, yet cannot grasp the whole without
understanding the parts---introduced a recursive structure into the
theory of meaning that had no formal counterpart until the recursive
type theories of the twentieth century.

Wilhelm Dilthey extended this programme by distinguishing sharply
between \emph{Erkl\"aren} (causal explanation, proper to the natural
sciences) and \emph{Verstehen} (empathetic understanding, proper to the
human sciences)~\cite{dilthey1883}.  For Dilthey, to understand a
cultural artefact was to re-experience the mental life of its creator
through the medium of shared cultural forms---a process he called
\emph{Nacherleben}.  The formal apparatus we develop in this chapter
offers a structural analogue: interpretation is the composition of the
receiver's prior ideas with the incoming signal, and the depth of the
resulting interpretant is bounded by the depths of its constituents.

Martin Heidegger radicalized the hermeneutic project further still.  In
\emph{Being and Time}~\cite{heidegger1927}, understanding is not a
cognitive act performed by a subject upon an object, but the fundamental
mode of human existence---\emph{Dasein} always already finds itself
within a horizon of pre-understanding (\emph{Vorverst\"andnis}).  Our
notion of a \emph{repertoire} formalizes precisely this pre-understanding:
it is the structured totality of ideas that a mind brings to any
encounter with the world.

Hans-Georg Gadamer's \emph{Truth and Method}~\cite{gadamer1960}
completed the twentieth-century hermeneutic arc by arguing that
understanding is always a \emph{fusion of horizons}
(\emph{Horizontverschmelzung}): the interpreter's horizon merges with
that of the text or signal.  It is this metaphor that our definition of
interpretation (Definition~\ref{def:interpretation}) renders
mathematically precise.  The formal treatment that follows is thus not a
departure from the hermeneutic tradition but its natural continuation:
the move from metaphor to theorem, from philosophical argument to
machine-verified proof.

% ── Introduction ──────────────────────────────────────────────────────

Every human mind is a repertoire: a stock of ideas accumulated through
enculturation.  When a signal arrives---a gesture, an utterance, a ritual
act---the receiver does not passively absorb it.  They \emph{interpret}
it by composing it with their existing ideas.  This chapter builds the
foundational structures on which the entire sequel rests.

In the language of Bourdieu~\cite{bourdieu1984}, a repertoire is the
cognitive component of \emph{habitus}---the durable, transposable
dispositions through which agents perceive and act in the social world.
In the language of Saussure~\cite{saussure1916}, each element of the
repertoire is a signified, and interpretation is the operation that maps
a signifier to a new signified by composing it with an existing one.  In
the language of Peirce~\cite{peirce1931}, interpretation is the
generation of an \emph{interpretant} from the encounter of a sign with a
prior sign.

The payoff in a signalling game is \emph{never} about the signal
itself---it is always about the interpretations the signal produces in
other minds.  This insight, already present in Gadamer's
\emph{Truth and Method}~\cite{gadamer1960}, receives its first fully
formal treatment here.

\medskip
\noindent\textbf{Chapter overview.}  We introduce four core concepts:
\emph{repertoire} (\S\ref{sec:ch1:core}), \emph{interpretation} and the
\emph{hermeneutic bound} (\S\ref{sec:ch1:hermeneutic-bound}), \emph{silence}
(\S\ref{sec:ch1:silence}), and \emph{cultural consensus}
(\S\ref{sec:ch1:consensus}).
We close with the \emph{telephone game} (\S\ref{sec:ch1:telephone}) and
structural invariants (\S\ref{sec:ch1:invariants}).  All sixteen theorems
are machine-verified in Lean~4 with zero \texttt{sorry} axioms.

\subsection*{Historical Context: From Schleiermacher to Formal Models}

The idea that interpretation might admit mathematical treatment has a
longer pedigree than is commonly supposed.  Charles Sanders Peirce, who
was both a logician and a semiotician, repeatedly sought algebraic
formulations of sign relations in the 1880s and 1890s, though he never
achieved a complete axiomatization~\cite{peirce1931}.  In the
twentieth century, structuralist anthropology---particularly the work
of Claude L\'evi-Strauss on mythical transformations~\cite{levistrauss1962}---came
tantalizingly close to a formal algebra of meaning, but lacked the
proof-theoretic infrastructure to verify its claims.

The resistance to formalizing interpretation has deep roots.  Dilthey's
\emph{Erkl\"aren}/\emph{Verstehen} distinction was read by many---including
Gadamer himself---as implying that the human sciences are fundamentally
incompatible with mathematical formalization.  If understanding requires
empathetic participation in another's mental life, then how could an
equation or a proof capture it?  The answer, as this chapter
demonstrates, is that formal models need not \emph{replace}
hermeneutic understanding; they can \emph{constrain} it.  The
Hermeneutic Bound (Theorem~\ref{thm:total-hermeneutic-bound}) does
not tell us \emph{what} a receiver will understand---it tells us what
they \emph{cannot} understand, given the structural limitations of
their repertoire.

Ideatic Depth Theory bridges the gap between continental philosophy and
formal methods by taking the phenomenological insight seriously---every
act of understanding is conditioned by prior understanding---and
expressing it in a language (Lean~4) where the consequences can be
mechanically verified.  The result is not a reduction of hermeneutics to
mathematics, but an enrichment of both: philosophy gains precision, and
mathematics gains a new domain of application.\footnote{For a related
programme bridging formal methods and humanistic inquiry, see the work on
computational hermeneutics by Benzm\"uller and Fuenmayor, who use
automated theorem provers to explore Leibniz's ontological arguments.}

% ── §1. Core Definitions ─────────────────────────────────────────────
\section{Core Definitions}
\label{sec:ch1:core}

We work within an ideatic space $(\IS, \compose, \void, \depth, \res)$
as established in the IDT Foundations.

\begin{definition}[Repertoire]
\label{def:repertoire}
A \emph{repertoire} is a finite list of ideas:
\[
  \Rep = [r_1, r_2, \ldots, r_n] \in \IS^{*}.
\]
In Bourdieu's terms, this is the \emph{habitus} made formal---the
totality of schemas through which an agent perceives and interprets the
social world.
\end{definition}

\begin{definition}[Interpretation]
\label{def:interpretation}
When a receiver with repertoire $\Rep = [r_1, \ldots, r_n]$ encounters
signal~$s$, they produce the list of interpretations
\[
  \operatorname{interpret}(\Rep, s) 
  = [r_1 \compose s,\; r_2 \compose s,\; \ldots,\; r_n \compose s].
\]
This is Gadamer's \emph{Horizontverschmelzung} (fusion of horizons):
each background idea merges with the incoming signal.
\end{definition}

In Lean~4, the definition is:

\begin{lstlisting}[caption={Interpretation in Lean~4},label=lst:interpret]
def interpret (rep : Repertoire I) (signal : I) : List I :=
  rep.map (fun r => IdeaticSpace.compose r signal)
\end{lstlisting}

\begin{definition}[Maximum Interpretation Depth]
\label{def:maxinterpdepth}
The \emph{maximum interpretation depth} of a list of ideas is
\[
  \operatorname{maxInterpDepth}([a_1, \ldots, a_n])
  = \max_{i} \depth(a_i).
\]
\end{definition}

\begin{definition}[Depth Sum]
\label{def:depthsum}
The \emph{depth sum} of a list of ideas is
$\operatorname{depthSum}([a_1, \ldots, a_n]) = \sum_i \depth(a_i)$.
\end{definition}

\begin{example}[A Navajo Sand Painting Ceremony]
\label{ex:navajo-sand-painting}
Consider a Navajo \emph{hataa{\l}ii} (healer) performing a sand painting
ceremony for a patient.  The healer's repertoire includes ideas drawn
from decades of apprenticeship: knowledge of the Holy People, the
symbolic meaning of each colour and cardinal direction, the narrative arc
of the relevant chant cycle, and the specific diagnostic signs exhibited
by the patient.  Let us denote these elements $r_1, r_2, \ldots, r_n$,
where the ordering reflects the sequence in which they were acquired
during apprenticeship---an ordering that matters, because the healer's
teacher transmitted them in a deliberate pedagogical sequence.

When the healer encounters a particular configuration of symptoms (the
signal~$s$), interpretation produces the list
$[r_1 \compose s, \ldots, r_n \compose s]$.  Each element represents a
different facet of the healer's understanding: $r_1 \compose s$ might be
the diagnostic interpretation (``the patient has been exposed to
lightning contamination''), while $r_k \compose s$ yields the
therapeutic plan (``perform the Male Shootingway chant'').  The depth
sum of the interpreted list is bounded by
$\operatorname{depthSum}(\Rep) + n \cdot \depth(s)$, reflecting the
fact that even a master healer's understanding is constrained by the
complexity of their training repertoire and the presenting symptoms.
\end{example}

\begin{example}[The Rosetta Stone]
\label{ex:rosetta-stone}
The decipherment of the Rosetta Stone by Jean-Fran\c{c}ois Champollion
in 1822 illustrates the role of repertoire depth in interpretation.
Champollion's repertoire included fluent Coptic (the descendant of
ancient Egyptian), knowledge of Greek, familiarity with the Ptolemaic
cartouche conventions, and Thomas Young's prior partial
decipherment.\footnote{See Andrew Robinson, \emph{Cracking the Egyptian
Code} (London: Thames \& Hudson, 2012).}  Each of these repertoire
elements had substantial depth---Coptic alone encoded centuries of
linguistic evolution.  When Champollion encountered the hieroglyphic
text (the signal~$s$), his interpretation
$\operatorname{interpret}(\Rep, s)$ produced readings that combined his
Coptic knowledge with the visual signs, yielding phonetic values that
no purely pictographic theory could have produced.  The maximum
interpretation depth was bounded by
$\operatorname{maxInterpDepth}(\Rep) + \depth(s)$: even Champollion's
genius could not extract more meaning than his repertoire and the
stone's inscriptions jointly permitted.
\end{example}

\subsection*{Discussion: Why Repertoires Are Lists, Not Sets}

A natural question arises: why model repertoires as \emph{lists} (finite
ordered sequences with possible repetitions) rather than \emph{sets}
(unordered collections without duplicates)?  Three considerations
motivate this choice.

First, \emph{ordering reflects biographical sequence}.  The ideas in a
mind are not interchangeable; they arrive in a specific order through
enculturation, and earlier ideas condition the reception of later ones.
A child who learns the concept of \emph{reciprocity} before the concept
of \emph{justice} inhabits a different cognitive space from one who
learns them in the reverse order, even if the eventual repertoire
contains the same ideas.\footnote{This point connects to Vygotsky's
zone of proximal development~\cite{vygotsky1978}: the order of
acquisition determines which ideas are accessible at each stage.}

Second, \emph{multiplicity captures emphasis}.  A repertoire may contain
the same idea multiple times---reflecting repeated exposure, ritual
reinforcement, or traumatic imprinting.  A soldier who has experienced
combat carries multiple copies of threat-recognition schemas; a devotee
who has chanted the same mantra ten thousand times has a repertoire in
which that mantra's associated ideas appear with high multiplicity.  The
depth sum of a list with repeated elements is strictly greater than that
of the corresponding set, correctly modelling the intensified
interpretive salience of frequently reinforced ideas.

Third, \emph{lists compose cleanly}.  The \texttt{List.map} operation
preserves length (Theorem~\ref{thm:conserves-count}) and distributes
over concatenation (Theorem~\ref{thm:division-of-labor}).  These
structural properties would not hold for sets under the image of a
non-injective map, since distinct repertoire elements might produce
identical interpretations that a set would collapse.

% ── §2. The Hermeneutic Bound ────────────────────────────────────────
\section{The Hermeneutic Bound}
\label{sec:ch1:hermeneutic-bound}

``You can only understand what you are prepared to understand.''
The following two theorems make this aphorism precise.

\begin{theorem}[Total Hermeneutic Bound]
\label{thm:total-hermeneutic-bound}
For any repertoire $\Rep$ and signal~$s$,
\[
  \operatorname{depthSum}(\operatorname{interpret}(\Rep, s))
  \;\leq\;
  \operatorname{depthSum}(\Rep) + |\Rep| \cdot \depth(s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By induction on $|\Rep|$.  The base case ($\Rep = []$) is trivial.
For $\Rep = r :: \Rep'$, we use the depth-composition bound
$\depth(r \compose s) \leq \depth(r) + \depth(s)$ and the
inductive hypothesis on~$\Rep'$.
The full proof is verified in \texttt{IDT\_Signal\_Ch01.lean},
theorem \texttt{total\_hermeneutic\_bound}.
\end{proof}

\begin{remark}[Mass Media]
This theorem explains why mass media (one signal, many receivers) has
bounded total interpretive impact per receiver.  A richer repertoire
absorbs more, but the signal's contribution scales only linearly with
audience size.
\end{remark}

\begin{theorem}[MaxDepth Interpretation Bound]
\label{thm:max-interp-depth}
For any repertoire $\Rep$ and signal~$s$,
\[
  \operatorname{maxInterpDepth}(\operatorname{interpret}(\Rep, s))
  \;\leq\;
  \operatorname{maxInterpDepth}(\Rep) + \depth(s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By induction on $|\Rep|$.  At each step, we bound
$\depth(r \compose s) \leq \depth(r) + \depth(s)$ and propagate
through the maximum.  See theorem \texttt{max\_interp\_depth\_bound}.
\end{proof}

\begin{remark}
Even the most sophisticated listener adds only $\depth(s)$ to their
best existing understanding.  No single signal can vault a receiver
beyond their current horizon by more than the signal's intrinsic
complexity.
\end{remark}

\subsection*{Implications of the Hermeneutic Bound}

The Hermeneutic Bound is, at first glance, an obvious constraint: of
course you cannot understand more than your background prepares you to
understand.  Yet its formal statement has consequences that are far from
obvious.

First, the bound is \emph{tight in the worst case}.  There exist ideatic
spaces in which equality holds: a signal of depth~$d$ composed with a
repertoire element of depth~$d'$ yields an interpretant of depth exactly
$d + d'$.  This means the bound cannot be improved without additional
assumptions on the structure of the ideatic space.

Second, the bound provides an \emph{information-theoretic} reading of
interpretation.  If we regard depth as an analogue of
Shannon entropy,\footnote{The analogy is imperfect but instructive.
Shannon entropy measures uncertainty in a probability distribution;
depth measures structural complexity of an idea.  Both are subadditive
under composition, and both are zero for the trivial element (the
uniform distribution and $\void$, respectively).} then the
Hermeneutic Bound says that interpretation is \emph{subadditive}: the
complexity of the output cannot exceed the sum of the complexities of
the inputs.  This is the formal analogue of the second law of
thermodynamics for meaning---interpretation cannot create complexity
\emph{ex nihilo}.

Third, and most surprisingly, the bound implies that \emph{pedagogy
has diminishing returns at fixed signal complexity}.  A teacher who
sends signals of bounded depth~$d$ can increase a student's maximum
interpretation depth by at most~$d$ per lesson, regardless of the
teacher's own sophistication.  To transmit ideas of depth~$D$, the
teacher must either send $D/d$ successive signals (a curriculum) or
send a single signal of depth~$D$ (which may be incomprehensible to a
novice).  This is the formal basis of scaffolding in educational
theory~\cite{vygotsky1978}.

% ── §3. Silence and the Unsaid ───────────────────────────────────────
\section{Silence and the Unsaid}
\label{sec:ch1:silence}

\begin{theorem}[Silence is Transparent]
\label{thm:silence-transparent}
The void signal leaves the mind unchanged:
\[
  \operatorname{interpret}(\Rep, \void) = \Rep.
\]
\end{theorem}

\begin{proof}[Proof sketch]
By induction on $\Rep$.  For each $r \in \Rep$, the void-right identity
$r \compose \void = r$ yields the result immediately.
See \texttt{silence\_is\_transparent}.
\end{proof}

\begin{remark}[Ritual Silence]
Quaker meetings, Buddhist \emph{noble silence}, the minute of silence
for the dead---all exploit this principle.  By saying nothing, the
receiver's existing framework remains undisturbed.  The void signal is
the formal foundation of taboo: what must not be said leaves the mind in
its current state.
\end{remark}

\begin{theorem}[The Blank Receiver]
\label{thm:blank-receiver}
A receiver whose only idea is $\void$ interprets any signal as the
signal itself:
\[
  \operatorname{interpret}([\void], s) = [s].
\]
\end{theorem}

\begin{proof}[Proof sketch]
Direct computation: $\void \compose s = s$ by the void-left identity.
See \texttt{blank\_receiver}.
\end{proof}

\begin{remark}[The Objective Observer]
The impossible ideal of the ``objective observer'' requires total
cultural amnesia: a repertoire consisting solely of~$\void$.  Only then
does interpretation reduce to transparent reception.
\end{remark}

\begin{theorem}[Negative Capability / Keats's Theorem]
\label{thm:negative-capability}
If $\void \in \Rep$, then $s \in \operatorname{interpret}(\Rep, s)$.
\end{theorem}

\begin{proof}[Proof sketch]
Since $\void \in \Rep$, the map sends $\void \mapsto \void \compose s = s$,
so $s$ appears in the output.
See \texttt{negative\_capability}.
\end{proof}

\begin{remark}
Keats described \emph{negative capability} as the capacity for ``being
in uncertainties, Mysteries, doubts, without any irritable reaching after
fact \& reason'' \cite{keats1817}.  In our framework, a mind retaining a
slot of pure receptivity ($\void$ in the repertoire) captures the signal
in its original form alongside culturally mediated versions.
\end{remark}

\begin{example}[Zen Buddhism's Beginner's Mind]
\label{ex:beginners-mind}
Shunryu Suzuki's celebrated dictum---``In the beginner's mind there are
many possibilities, but in the expert's there are few''~\cite{suzuki1970}---receives
a precise formulation in our framework.  The ``beginner's mind''
(\emph{shoshin}, \begin{CJK}{UTF8}{min}初心\end{CJK}) corresponds to a
repertoire in which $\void$ occupies a prominent position.  By
Theorem~\ref{thm:negative-capability}, such a repertoire guarantees
that the signal~$s$ appears unaltered among the interpretations.  The
``expert's mind,'' by contrast, may have displaced $\void$ entirely,
replacing it with highly specific schemas.  Every interpretation is
then filtered through these schemas, and the raw signal is lost.

The Zen practice of \emph{zazen} (sitting meditation) can be understood
as an attempt to restore $\void$ to a repertoire that has become
saturated with expert knowledge.  The practitioner cultivates a
deliberate emptiness---not the absence of ideas, but the presence of
$\void$ alongside existing ideas---so that new signals can be received
in their unmediated form.\footnote{This reading aligns with Nishitani's
account of \emph{sh\=unyat\=a} (emptiness) as a positive presence
rather than mere negation; see Keiji Nishitani, \emph{Religion and
Nothingness}, trans.\ Jan Van Bragt (Berkeley: University of California
Press, 1982).}
\end{example}

\begin{remark}[Connection to Machine Learning]
\label{rem:ml-void}
The void element in the repertoire bears a striking structural
resemblance to zero-initialization and residual connections in deep
neural networks.  In a residual network~\cite{he2016resnet}, the output
of a layer is $f(x) + x$: the input is preserved alongside the
transformed version, just as $\void \compose s = s$ preserves the signal
alongside culturally mediated interpretations.  The skip connection
ensures that the ``raw signal'' propagates through the network
unattenuated---a computational analogue of negative capability.  This
suggests that the architectural innovations of modern deep learning may
be independently rediscovering structural principles that the
hermeneutic tradition identified centuries ago.
\end{remark}

% ── §4. Cultural Consensus ───────────────────────────────────────────
\section{Cultural Consensus}
\label{sec:ch1:consensus}

The next three theorems establish that resonance in backgrounds implies
resonance in interpretations---the structural foundation of cultural
consensus.

\begin{theorem}[Durkheimian Consensus]
\label{thm:durkheimian-consensus}
If $r_1 \res r_2$ and both receive signal~$s$, then
$r_1 \compose s \;\res\; r_2 \compose s$.
\end{theorem}

\begin{proof}[Proof sketch]
Direct application of the resonance-compose-right lemma from IDT
Foundations.  See \texttt{durkheimian\_consensus}.
\end{proof}

\begin{remark}
Two members of the same culture (shared myths, shared categories)
\emph{always} produce resonant interpretations of the same event.
Cultural consensus is a structural consequence of shared
background---Durkheim's \emph{conscience collective}~\cite{durkheim1893}.
\end{remark}

\begin{theorem}[L\'evi-Strauss's Mythical Transformation]
\label{thm:mythical-transformation}
For any receiver~$r$, if $s_1 \res s_2$ then
$r \compose s_1 \;\res\; r \compose s_2$.
\end{theorem}

\begin{proof}[Proof sketch]
By resonance-compose-left.  See \texttt{mythical\_transformation}.
\end{proof}

\begin{remark}
Ritual repetition-with-variation works because the participant's
invariant background ensures that structurally similar performances
produce resonant experiences.  L\'evi-Strauss's mythical
``transformations'' are perceived as ``the same story told differently''
precisely because the listener's background resonates with both
variants~\cite{levistrauss1962}.
\end{remark}

\begin{theorem}[Intercultural Solidarity / Turner's Communitas]
\label{thm:turners-communitas}
If $r_1 \res r_2$ and $s_1 \res s_2$, then
\[
  r_1 \compose s_1 \;\res\; r_2 \compose s_2.
\]
\end{theorem}

\begin{proof}[Proof sketch]
By the full resonance-compose lemma.  See \texttt{turners\_communitas}.
\end{proof}

\begin{remark}
Members of allied cultures witnessing similar rituals produce resonant
interpretations.  This is the formal basis of what Turner called
\emph{communitas}~\cite{turner1969}: the spontaneous solidarity that
arises when structurally similar participants undergo similar experiences.
\end{remark}

% ── §5. The Telephone Game ───────────────────────────────────────────
\section{The Telephone Game: Cultural Transmission Chains}
\label{sec:ch1:telephone}

\begin{theorem}[Bartlett's Serial Reproduction]
\label{thm:bartlett}
Two-stage interpretation through different minds equals single-stage
interpretation through the composed mind:
\[
  r_B \compose (r_A \compose s)
  \;=\;
  (r_B \compose r_A) \compose s.
\]
\end{theorem}

\begin{proof}[Proof sketch]
This is the associativity axiom of the ideatic space, applied to the
triple $(r_B, r_A, s)$.
See \texttt{bartletts\_serial\_reproduction}.
\end{proof}

\begin{remark}
Bartlett's (1932) serial reproduction experiments~\cite{bartlett1932}
showed that stories change predictably through chains of retelling.  This
theorem shows \emph{why}: the chain of interpreters is structurally
equivalent to a single composite interpreter.  Oral tradition scholars
can study the ``net effect'' of a transmission chain without tracking
each individual retelling.
\end{remark}

\begin{theorem}[Telephone Depth Bound]
\label{thm:telephone-depth}
The depth after two-stage interpretation satisfies
\[
  \depth(r_B \compose (r_A \compose s))
  \;\leq\;
  \depth(r_B) + \depth(r_A) + \depth(s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
Rewrite using associativity, then apply depth-compose-le twice.
See \texttt{telephone\_depth}.
\end{proof}

\begin{theorem}[Resonant Transmission Lines]
\label{thm:resonant-transmission}
If $r_{A_1} \res r_{A_2}$ and $r_{B_1} \res r_{B_2}$, then
\[
  r_{B_1} \compose (r_{A_1} \compose s)
  \;\res\;
  r_{B_2} \compose (r_{A_2} \compose s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
Apply associativity to both sides, then use resonance-compose twice.
See \texttt{resonant\_transmission\_lines}.
\end{proof}

\begin{remark}
Independent lines of ritual transmission (e.g., different griots in West
Africa, different monastic orders transmitting the same texts) produce
resonant outcomes if corresponding specialists resonate.  Ritual
orthodoxy can be maintained across geographically separated communities
without central authority.
\end{remark}

\begin{example}[The Homeric Tradition]
\label{ex:homeric-tradition}
The transmission of Homer's \emph{Iliad} across centuries of oral
tradition provides a vivid illustration of
Theorem~\ref{thm:resonant-transmission}.  The Homeric \emph{rhapsodes}
(professional reciters) formed parallel transmission lines: different
performers in different cities---Athens, Chios, Smyrna---each received
the epic from their own teacher and transmitted it to their own
students.\footnote{See Albert B.\ Lord, \emph{The Singer of Tales}
(Cambridge, MA: Harvard University Press, 1960), for the definitive
account of oral-formulaic composition and transmission.}  Let $r_{A_1}$
and $r_{A_2}$ denote the repertoires of two first-generation rhapsodes
trained by the same master, and $r_{B_1}$, $r_{B_2}$ their respective
students.  If corresponding performers resonate ($r_{A_1} \res r_{A_2}$
and $r_{B_1} \res r_{B_2}$), then the theorem guarantees that the
versions reaching the second generation are resonant---structurally
``the same poem''---despite the absence of a written text to enforce
uniformity.  This explains a long-standing puzzle in Homeric scholarship:
how could an oral tradition of such length and complexity remain
coherent across centuries without writing?  The answer is structural:
resonance in the transmission chain is sufficient.
\end{example}

\subsection*{Discussion: Cultural Transmission as Algebraic Morphism}

The telephone-game theorems invite a categorical perspective.  If we
regard the ideatic space $(\IS, \compose)$ as a monoid, then each
element $r \in \IS$ defines a left-multiplication map
$\lambda_r : \IS \to \IS$ by $\lambda_r(s) = r \compose s$.
Bartlett's serial reproduction theorem
(Theorem~\ref{thm:bartlett}) is then the statement that
$\lambda_{r_B} \circ \lambda_{r_A} = \lambda_{r_B \compose r_A}$: the
composition of transmission maps equals the transmission map of the
composed element.  In categorical language, the assignment
$r \mapsto \lambda_r$ is a monoid homomorphism from $(\IS, \compose)$
to $(\operatorname{End}(\IS), \circ)$.

This observation has a profound consequence: the study of cultural
transmission chains reduces to the study of a single algebraic object,
the monoid $(\IS, \compose)$.  Every question about what happens when a
signal passes through a sequence of interpreters can be answered by
examining the product of the corresponding elements.  The resonant
transmission theorem further ensures that this reduction is compatible
with the equivalence relation $\res$: resonance classes are preserved
under the monoid action.\footnote{In categorical terms, resonance
defines a congruence on the monoid, and the quotient
$\IS / {\res}$ inherits a monoid structure.}

% ── §6. Structural Invariants ────────────────────────────────────────
\section{Structural Invariants of Interpretation}
\label{sec:ch1:invariants}

\begin{theorem}[Void is Unpersuasive]
\label{thm:void-unpersuasive}
$\depth(r \compose \void) = \depth(r)$.  Silence contributes zero
marginal depth.
\end{theorem}

\begin{proof}[Proof sketch]
By the depth-compose-void identity.  See \texttt{void\_is\_unpersuasive}.
\end{proof}

\begin{remark}[Game-theoretic significance]
The null message (babbling) contributes zero information.  This is
\emph{why} babbling equilibria always exist: silence costs nothing and
changes nothing.
\end{remark}

\begin{theorem}[Interpretation Conserves Count]
\label{thm:conserves-count}
$|\operatorname{interpret}(\Rep, s)| = |\Rep|$.
\end{theorem}

\begin{proof}[Proof sketch]
By the length-preserving property of \texttt{List.map}.
See \texttt{interpretation\_conserves\_count}.
\end{proof}

\begin{theorem}[Durkheim's Division of Labor]
\label{thm:division-of-labor}
$\operatorname{interpret}(\Rep_1 \mathbin{+\!+} \Rep_2, s)
 = \operatorname{interpret}(\Rep_1, s) \mathbin{+\!+}
   \operatorname{interpret}(\Rep_2, s)$.
\end{theorem}

\begin{proof}[Proof sketch]
By \texttt{List.map\_append}.  See \texttt{durkheims\_division\_of\_labor}.
\end{proof}

\begin{remark}
A polymath's interpretation equals the union of specialist
interpretations.  Interdisciplinary understanding is \emph{not}
synergistic---it is the disjoint union of specialist understandings.
This formalizes Durkheim's cognitive division of labor~\cite{durkheim1893}.
\end{remark}

\begin{theorem}[Silence Preserves DepthSum]
\label{thm:silence-depthsum}
$\operatorname{depthSum}(\operatorname{interpret}(\Rep, \void))
 = \operatorname{depthSum}(\Rep)$.
\end{theorem}

\begin{proof}[Proof sketch]
Follows immediately from Theorem~\ref{thm:silence-transparent}.
See \texttt{silence\_preserves\_depthSum}.
\end{proof}

\begin{theorem}[Three-Stage Telephone]
\label{thm:three-stage}
\[
  r_C \compose (r_B \compose (r_A \compose s))
  \;=\;
  ((r_C \compose r_B) \compose r_A) \compose s.
\]
\end{theorem}

\begin{proof}[Proof sketch]
Three applications of associativity.
See \texttt{three\_stage\_telephone}.
\end{proof}

\begin{remark}
Arbitrarily long chains of cultural transmission collapse into a single
composite interpreter.  This is why we can meaningfully speak of ``the
Western tradition'' as a single interpretive lens, despite it being the
product of thousands of individual acts of interpretation.
\end{remark}

\begin{example}[The Game of ``Telephone'' in Cross-Cultural Settings]
\label{ex:telephone-cross-cultural}
Empirical studies of the telephone game (also known as ``Chinese
whispers'' or \emph{stille Post}) reveal striking cross-cultural
variation in transmission fidelity.  Kashima~\cite{kashima2000}
demonstrated that culturally shared stereotypes are transmitted with
higher fidelity than counter-stereotypical information, a finding that
our framework predicts: if the repertoire elements $r_A, r_B, r_C$
in a transmission chain all contain a common idea~$r^*$ (a shared
stereotype), then the component $r^* \compose s$ of each
interpretation is preserved at every stage.  The three-stage telephone
theorem guarantees that $r_C \compose (r_B \compose (r_A \compose s))
= ((r_C \compose r_B) \compose r_A) \compose s$, and the shared
component~$r^*$ acts as a fixed point under this iterated composition.

Conversely, information that no member of the chain has in their
repertoire---culturally alien content---degrades rapidly, because
the composition $r_i \compose s$ distorts $s$ in a direction determined
by $r_i$, and successive distortions compound.  This is the formal
basis of Bartlett's~\cite{bartlett1932} classic finding that stories
undergo ``conventionalization'' during serial reproduction: details
foreign to the listeners' cultural repertoire are systematically
replaced by familiar ones.
\end{example}

\begin{remark}[Open Questions]
\label{rem:open-questions}
Several natural questions remain beyond the scope of this chapter's
theorems.  First, what happens if the composition operation $\compose$
is \emph{not} associative?  Non-associative interpretation would mean
that the order in which a chain of interpreters is bracketed matters:
$(r_C \compose r_B) \compose r_A \neq r_C \compose (r_B \compose r_A)$.
Such structures arise naturally in the study of Lie algebras and
quasigroups, and may model situations where the \emph{context} of
interpretation (who spoke before whom) affects the outcome in ways not
captured by the signal content alone.

Second, is there an analogue of the Hermeneutic Bound for
\emph{probabilistic} interpretation, where the receiver samples from a
distribution over repertoire elements rather than applying each one
deterministically?  Such a ``stochastic hermeneutics'' would connect
IDT to Bayesian models of cognition~\cite{tenenbaum2011} and might
yield tighter bounds in the average case.

Third, can the resonance relation $\res$ be given a
\emph{metric} refinement, so that we can speak not only of whether two
interpretations resonate but of \emph{how closely} they resonate?  A
metric hermeneutics would enable quantitative predictions about the
degree of cultural consensus as a function of repertoire overlap.
\end{remark}

% ── Summary ──────────────────────────────────────────────────────────
\section*{Chapter Summary}

This chapter established the foundational apparatus of the signalling-game
framework within IDT\@.  We introduced \emph{repertoires} as formal models
of culturally constituted minds, and \emph{interpretation} as
compositional fusion of horizons.  The \emph{Total Hermeneutic Bound}
(Theorem~\ref{thm:total-hermeneutic-bound}) shows that no signal can
inject more complexity into a mind than the signal's depth times the
repertoire's size.  \emph{Silence is Transparent}
(Theorem~\ref{thm:silence-transparent}) provides the formal foundation
for babbling equilibria.  The \emph{Durkheimian Consensus}
(Theorem~\ref{thm:durkheimian-consensus}) and \emph{Turner's Communitas}
(Theorem~\ref{thm:turners-communitas}) establish that resonant
backgrounds guarantee resonant interpretations, providing the formal
basis of cultural consensus.  The \emph{Telephone Game} theorems
(Theorems~\ref{thm:bartlett}--\ref{thm:resonant-transmission}) show that
sequential interpretation chains collapse into single composite
interpreters via associativity.

All sixteen theorems are verified in
\texttt{FormalAnthropology/IDT\_Signal\_Ch01.lean}.
