% ══════════════════════════════════════════════════════════════════════
% Chapter 2: The Signalling Problem
% ══════════════════════════════════════════════════════════════════════

\chapter{The Signalling Problem}
\label{ch:signalling}

\epigraph{``Words are also actions, and actions are a kind of words.''}{Ralph Waldo Emerson}

% ── Introduction ──────────────────────────────────────────────────────

The study of strategic communication has a distinguished lineage in
philosophy, economics, and the social sciences.  David Lewis's
\emph{Convention}~\cite{lewis1969} established that signalling systems
can arise purely from coordination, without any prior agreement or
external enforcement---a result that shattered the prevailing view that
meaning must be stipulated by authority.\footnote{Lewis's central
insight was that a ``convention'' is a regularity in behaviour, sustained
by mutual expectations and mutual preferences, that solves a
\emph{coordination problem}.  Signalling conventions are the paradigm
case: sender and receiver have no conflict of interest, only a shared
need to coordinate on which signal maps to which meaning.}  Lewis's
framework, however, assumed common knowledge of a shared signalling
code, leaving open the question of how such codes originate.

Brian Skyrms's \emph{Signals: Evolution, Learning, and
Information}~\cite{skyrms2010} took up precisely this question, showing
through evolutionary game theory that signalling conventions can emerge
from initially random behaviour through reinforcement learning and
natural selection.  Skyrms demonstrated that even the simplest learning
dynamics---such as Roth--Erev reinforcement---converge to signalling
equilibria with high probability, suggesting that the capacity for
meaningful communication requires no cognitive sophistication beyond the
ability to repeat rewarded behaviour.

Crawford and Sobel's~\cite{crawford1982} cheap talk model introduced
the possibility of \emph{conflict} between sender and receiver.  In
their framework, a sender with private information transmits a
costless message to a receiver who then takes an action affecting
both parties' payoffs.  The central result is that the amount of
information transmitted in equilibrium decreases as the divergence
of interests increases.  When interests are perfectly aligned, full
revelation is possible; when they are opposed, only a ``babbling
equilibrium'' survives in which the signal carries no information
at all.\footnote{Cho and Kreps~\cite{cho1987} refined the analysis
of signalling games by introducing the \emph{Intuitive Criterion},
which eliminates implausible equilibria by asking whether a deviation
could be ``reasonably'' attributed to a particular sender type.
Their refinement has been enormously influential in economics but
assumes that receivers can reason counterfactually about sender
types---an assumption that IDT replaces with the compositional
structure of repertoires.}

The framework developed in this chapter departs from all of these
predecessors in a single decisive respect: \emph{payoffs depend on
interpretations, not on signals or actions}.  In Crawford and Sobel's
model, the receiver takes an action (e.g., choosing a policy) and
payoffs depend on that action and the state of the world.  In IDT,
there is no ``action'' separate from interpretation---the receiver's
response \emph{is} the composition of the signal with their repertoire,
and the sender's payoff depends on the resulting list of composed
ideas.  This shift has profound consequences.  It means that the
same signal, sent to two receivers with different repertoires, can
yield entirely different payoffs for the sender---not because the
receivers ``chose'' different responses, but because their existing
conceptual structures composed differently with the signal.

In every human society, actors choose signals strategically: a politician
picks rhetoric, a suitor offers gifts, a priest designs liturgy.  The
sender's fundamental problem is that they cannot control how receivers
will \emph{interpret} the signal---each receiver composes the signal with
their own repertoire.

This chapter formalises the core game-theoretic structure that governs
strategic communication.  Following Spence~\cite{spence1973} and
Crawford \& Sobel~\cite{crawford1982}, we model the signalling problem
as a game in which a sender chooses a signal, a receiver interprets it
through their repertoire, and both receive payoffs from the resulting
interpretations.  The crucial departure from classical game theory is
that payoffs depend not on the signal itself but on what receivers
\emph{make of it}---how it composes with their existing repertoire of
ideas.

\medskip
\noindent\textbf{Chapter overview.}  We define the \emph{signalling game}
(\S\ref{sec:ch2:game}), establish the \emph{babbling equilibrium}
(\S\ref{sec:ch2:babbling}), analyse \emph{audience structure}
(\S\ref{sec:ch2:audience}), prove \emph{depth bounds} on outcomes
(\S\ref{sec:ch2:depth}), study \emph{resonance in game outcomes}
(\S\ref{sec:ch2:resonance}), and explore \emph{strategic properties}
(\S\ref{sec:ch2:strategic}).  All eighteen theorems are machine-verified.

\subsection*{Historical Context: From Animal Signals to Game Theory}

The application of game-theoretic reasoning to communication long
predates its formalisation in economics.  Zahavi's \emph{handicap
principle}~\cite{zahavi1975} proposed that animal signals must be
\emph{costly} to be honest: the peacock's tail is reliable precisely
because it is expensive to produce and maintain, so only genuinely
fit males can afford it.  Maynard Smith and Harper~\cite{maynardsmith2003}
systematised this insight in \emph{Animal Signals}, distinguishing
between ``assessment signals'' (whose reliability is guaranteed by
physical constraints) and ``conventional signals'' (whose reliability
depends on equilibrium maintenance).

Anthropologists were quick to adopt these models.  Irons~\cite{irons2001}
applied the handicap principle to religious ritual, arguing that costly
ceremonies serve as honest signals of commitment to the group.  Sosis
and Alcorta~\cite{sosis2003} extended this analysis to explain the
persistence of apparently irrational religious behaviours as adaptive
signalling mechanisms.  Henrich~\cite{henrich2009} argued that
``credibility-enhancing displays''---actions that are costly and
consistent with professed beliefs---are the primary mechanism by which
cultural learners identify reliable models to imitate.

The IDT framework subsumes these approaches by shifting attention from
the \emph{cost} of the signal to the \emph{structure} of the receiver's
repertoire.  Whether a signal is ``honest'' in the Zahavian sense
depends not on its production cost but on how it composes with what
receivers already believe.  A cheap signal can be perfectly informative
to a receiver whose repertoire is aligned with the sender's intentions,
while an expensive signal can be entirely uninformative to a receiver
whose conceptual apparatus is orthogonal to the sender's meaning.
This reframing dissolves the sharp distinction between ``cheap talk''
and ``costly signalling'' that has organised the literature since
Crawford and Sobel.\footnote{The dissolution is not merely
philosophical.  Formally, cost enters IDT through the depth of the
signal: a ``costly'' signal is one of high depth, and
Theorem~\ref{thm:outcome-depth-bound} shows that the effect of depth
is linearly bounded by the receiver's existing depth.  Cost without
receptivity is wasted.}

% ── §1. The Signalling Game ──────────────────────────────────────────
\section{The Signalling Game}
\label{sec:ch2:game}

\begin{definition}[Signalling Game]
\label{def:signalling-game}
A \emph{signalling game} $G = (\Rep, \Pay_S, \Pay_R)$ consists of:
\begin{itemize}
  \item A receiver repertoire $\Rep \in \IS^*$;
  \item A sender payoff function $\Pay_S : \IS^* \to \mathbb{N}$;
  \item A receiver payoff function $\Pay_R : \IS^* \to \mathbb{N}$.
\end{itemize}
The payoff depends on the \emph{interpretation} (the list of composed
ideas), not the signal directly.
\end{definition}

\begin{lstlisting}[caption={SignallingGame in Lean~4},label=lst:signalling-game]
structure SignallingGame (I : Type*) [IdeaticSpace I] where
  repertoire : Repertoire I
  sender_payoff : List I -> Nat
  receiver_payoff : List I -> Nat
\end{lstlisting}

\begin{definition}[Outcome]
\label{def:outcome}
The \emph{outcome} of sending signal~$s$ in game~$G$ is
\[
  \operatorname{outcome}(G, s) = \operatorname{interpret}(G.\Rep, s).
\]
\end{definition}

\begin{definition}[Optimal Signal]
\label{def:optimal-signal}
A signal~$s$ is \emph{optimal} for the sender if no alternative yields
strictly higher payoff:
\[
  \forall s'.\; \Pay_S(\operatorname{outcome}(G, s'))
  \leq \Pay_S(\operatorname{outcome}(G, s)).
\]
\end{definition}

\begin{definition}[Constant Sender Payoff]
\label{def:constant-payoff}
A game has \emph{constant sender payoff} if
$\Pay_S(\mathbf{x}) = \Pay_S(\mathbf{y})$ for all interpretation
lists~$\mathbf{x}, \mathbf{y}$.
\end{definition}

% ── §2. Babbling Equilibrium ─────────────────────────────────────────
\section{Babbling Equilibrium}
\label{sec:ch2:babbling}

\begin{theorem}[Babbling Outcome]
\label{thm:babbling-outcome}
Sending void produces the original repertoire unchanged:
\[
  \operatorname{outcome}(G, \void) = G.\Rep.
\]
\end{theorem}

\begin{proof}[Proof sketch]
Unfold the definitions and apply Theorem~\ref{thm:silence-transparent}
from Chapter~1.  See \texttt{babbling\_outcome}.
\end{proof}

\begin{remark}
The ``babbling equilibrium'' of Crawford \& Sobel~\cite{crawford1982}
exists in every signalling game because silence is transparent---the
receiver's mind is left exactly as it was.
\end{remark}

\begin{theorem}[Constant Payoff Babbling Optimality]
\label{thm:constant-payoff}
If sender payoff is constant, every signal is optimal---including void.
\end{theorem}

\begin{proof}[Proof sketch]
If $\Pay_S$ is constant, $\Pay_S(\operatorname{outcome}(G,s'))
= \Pay_S(\operatorname{outcome}(G,s))$ for all $s,s'$, so the
optimality condition holds trivially.  See \texttt{constant\_payoff\_optimal}.
\end{proof}

\begin{remark}
When the sender genuinely does not care what the receiver thinks,
information transmission collapses entirely.  This is the cheapest of
cheap talk.
\end{remark}

\begin{theorem}[Outcome Length Invariance]
\label{thm:outcome-length}
Every signal produces the same number of interpretations:
\[
  |\operatorname{outcome}(G, s)| = |G.\Rep|.
\]
\end{theorem}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:conserves-count}.  See \texttt{outcome\_length}.
\end{proof}

\begin{remark}
You cannot make someone think \emph{more} thoughts by choosing a
cleverer signal.  The number of schemas activated is a fixed property of
the receiver, not the sender.
\end{remark}

\begin{theorem}[Singleton Repertoire Outcome]
\label{thm:singleton-outcome}
A receiver with a single idea~$r$ produces $[r \compose s]$.
\end{theorem}

\begin{proof}[Proof sketch]
Direct computation.  See \texttt{singleton\_outcome}.
\end{proof}

\begin{remark}
The fanatic (single-idea mind) has a perfectly predictable interpretation
of any signal.  Totalitarian receivers are, paradoxically, the easiest
to communicate with.
\end{remark}

\begin{example}[Political Rhetoric and the Single-Issue Voter]
Consider a political campaign targeting single-issue voters.  A voter
whose entire repertoire consists of a single idea---say, $r = \text{``gun
rights''}$---will interpret \emph{every} campaign signal through that
lens.  A speech about healthcare becomes $r \compose s_{\text{health}}$;
a speech about foreign policy becomes $r \compose s_{\text{foreign}}$.
By Theorem~\ref{thm:singleton-outcome}, each speech produces exactly one
interpretation, and that interpretation is entirely determined by the
voter's fixed idea composed with the signal.  The campaign strategist
thus faces a simple optimisation problem: choose the signal~$s$ that
maximises $\Pay_S([r \compose s])$.  In practice, this means framing
\emph{every} issue in terms of gun rights, regardless of the actual
topic---a strategy familiar to any observer of American politics.
\end{example}

\begin{example}[The Azan (Islamic Call to Prayer)]
\label{ex:azan}
The muezzin's call to prayer (azan) is a signal broadcast to an entire
city.  Its \emph{content} is fixed---the same Arabic phrases, five times
daily---but its \emph{interpretations} vary dramatically across
receivers.  For the devout Muslim, the azan composes with a deep
repertoire of theological ideas, producing an interpretation that
activates obligations, memories of communal worship, and eschatological
awareness.  For the tourist, the azan composes with a shallow repertoire
of aesthetic or orientalist ideas, producing an interpretation centred on
exoticism or architectural appreciation.  For the neighbouring resident
of a different faith, the repertoire may contain ideas of territorial
assertion or noise pollution, yielding an entirely different composed
idea.\footnote{This example illustrates why Theorem~\ref{thm:outcome-length}
is significant: the azan produces exactly $|G.\Rep|$ interpretations,
one per schema in the receiver's repertoire.  The \emph{number} of
thoughts provoked is a property of the receiver, not the signal.}

The IDT framework captures this without resorting to ``context'' as an
unexplained primitive.  The differences in interpretation are fully
determined by the differences in repertoire, and the sender (the
muezzin, or the institution that mandates the call) can reason about
these differences through the audience structure theorems of
\S\ref{sec:ch2:audience}.
\end{example}

\subsection*{Discussion: Why Payoffs Depend on Interpretations}

The decision to make payoffs depend on \emph{interpretations} rather
than \emph{actions} is the central modelling choice of this chapter, and
it warrants explicit justification.  In classical game theory, a receiver
observes a signal and then ``chooses an action'' from some action set.
Payoffs depend on the action chosen and possibly on the state of the
world.  This framework assumes that interpretation and action are
separate: the receiver first \emph{understands} the signal, then
\emph{decides} what to do.

IDT collapses this distinction.  The receiver's ``action'' \emph{is}
their interpretation---the composition $r \compose s$ of the signal with
their repertoire.  There is no separate decision stage because the
receiver does not ``choose'' how to interpret; interpretation is
determined by the compositional structure of their existing ideas.  This
modelling choice is motivated by the ethnographic observation that most
cultural communication does not involve deliberate choice at the point
of reception.  When a congregant hears a sermon, they do not first
decode the words and then select a response; their response \emph{is}
the composition of the sermon with their existing beliefs, memories,
and conceptual categories.  The payoff to the preacher depends on what
the congregant \emph{thinks}, not on what they ``do'' in any
behaviourist sense.

% ── §3. Audience Structure ───────────────────────────────────────────
\section{Audience Structure}
\label{sec:ch2:audience}

\begin{theorem}[Merged Audience]
\label{thm:merged-audience}
Signalling to a combined audience equals the concatenation of separate
outcomes:
\[
  \operatorname{outcome}(\Rep_1 \mathbin{+\!+} \Rep_2, s)
  = \operatorname{outcome}(\Rep_1, s) \mathbin{+\!+}
    \operatorname{outcome}(\Rep_2, s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By \texttt{List.map\_append}.  See \texttt{merged\_audience\_outcome}.
\end{proof}

\begin{remark}
A political speech to a diverse audience (workers $\cup$ elites) produces
interpretations that are the disjoint union of what each subgroup would
have produced separately.  There is no ``synergistic'' interpretation
effect from mere co-presence.
\end{remark}

\begin{theorem}[Empty Audience]
\label{thm:empty-audience}
Signalling to no one produces nothing:
$\operatorname{outcome}([], s) = []$.
\end{theorem}

\begin{proof}[Proof sketch]
By definition.  See \texttt{empty\_audience}.
\end{proof}

\begin{remark}
If nobody is there to interpret, there are no interpretations.  A signal
without a receiver is not a communicative act---formalising Wittgenstein's
point that meaning requires use~\cite{wittgenstein1953}.
\end{remark}

\begin{theorem}[Blank Audience Transparency]
\label{thm:blank-audience}
A receiver whose repertoire is all void interprets any signal as
$n$~copies of the signal itself:
\[
  \operatorname{outcome}(\operatorname{replicate}(n, \void), s)
  = \operatorname{replicate}(n, s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By induction on~$n$, using $\void \compose s = s$ at each step.
See \texttt{blank\_audience}.
\end{proof}

\begin{remark}[Connection to Information Theory]
\label{rem:info-theory}
The audience structure theorems have a natural information-theoretic
interpretation.  The merged audience theorem
(Theorem~\ref{thm:merged-audience}) states that interpretation
distributes over concatenation of audiences.  In Shannon's
framework~\cite{shannon1948}, this corresponds to the statement that
sending the same codeword through two independent channels produces
independent outputs---the channels do not interact.  The blank audience
theorem (Theorem~\ref{thm:blank-audience}) identifies a special case:
an audience of void ideas is an \emph{identity channel} that transmits
the signal unchanged.  In information-theoretic terms, the blank
audience has maximum channel capacity---it introduces no noise, no
distortion, and no loss.  Every bit of the sender's signal passes
through unaltered.

This analogy should not be pushed too far.  Unlike Shannon channels,
IDT composition is not stochastic; it is a deterministic algebraic
operation.  But the structural parallel illuminates why the audience
theorems take the form they do: they describe how the ``channel''
(the receiver's repertoire) transforms the ``message'' (the signal).
\end{remark}

\begin{example}[Cargo Cults as Blank Audience Phenomena]
\label{ex:cargo-cults}
The cargo cults of Melanesia provide a striking illustration of
near-blank audience reception.\footnote{See Worsley~\cite{worsley1957}
for the classic anthropological treatment, and Lindstrom~\cite{lindstrom1993}
for a critical reassessment of the ``cargo cult'' concept itself.}  When
Pacific Islanders first encountered Allied military technology during
World War II, many communities had no existing repertoire of ideas about
industrial production, global logistics, or military strategy.  Their
repertoire, with respect to these signals, was close to void.

By Theorem~\ref{thm:blank-audience}, a near-void repertoire produces
interpretations close to the signal itself.  The Melanesian observers
thus ``received'' the signals---the arrival of ships, the construction
of airstrips, the delivery of cargo---more or less as raw phenomena,
unmediated by the conceptual apparatus that would have led a Western
observer to interpret them as instances of industrial logistics.  The
ritual reproduction of airstrips and control towers was not
``irrational''; it was the natural consequence of composing novel
signals with a repertoire that lacked the relevant compositional
structure.  The signal passed through nearly unchanged, and the
resulting interpretation was the signal itself.
\end{example}

% ── §4. Depth Bounds on Outcomes ─────────────────────────────────────
\section{Depth Bounds on Outcomes}
\label{sec:ch2:depth}

\begin{theorem}[Outcome Depth Bound]
\label{thm:outcome-depth-bound}
\[
  \operatorname{depthSum}(\operatorname{outcome}(G, s))
  \leq \operatorname{depthSum}(G.\Rep) + |G.\Rep| \cdot \depth(s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
Direct application of the Total Hermeneutic Bound
(Theorem~\ref{thm:total-hermeneutic-bound}).
See \texttt{outcome\_depth\_bound}.
\end{proof}

\begin{remark}
The sender's capacity to inject complexity into the receiver's mind is
linearly bounded.  This is why propaganda targets breadth (many receivers)
rather than depth (one deep signal).
\end{remark}

\begin{theorem}[Void Signal Zero Cost]
\label{thm:void-zero-cost}
The void signal adds zero total depth to the outcome:
\[
  \operatorname{depthSum}(\operatorname{outcome}(G, \void))
  = \operatorname{depthSum}(G.\Rep).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:silence-depthsum}.  See \texttt{void\_signal\_zero\_cost}.
\end{proof}

% ── §5. Resonance in Game Outcomes ───────────────────────────────────
\section{Resonance in Game Outcomes}
\label{sec:ch2:resonance}

\begin{theorem}[Resonant Repertoire Consensus]
\label{thm:resonant-repertoire}
If all elements in the repertoire resonate with a fixed idea~$a$, then
all interpretations of any signal also resonate with $a \compose s$.
\end{theorem}

\begin{proof}[Proof sketch]
For each $r \in \Rep$ with $r \res a$, by Theorem~\ref{thm:durkheimian-consensus}
we get $r \compose s \res a \compose s$.
See \texttt{resonant\_repertoire\_consensus}.
\end{proof}

\begin{remark}
A culturally homogeneous audience (everyone resonates with the same
archetype) produces interpretations that all resonate with that
archetype's interpretation.  Cultural consensus is self-reinforcing.
\end{remark}

\begin{theorem}[Resonant Signals Produce Resonant Outcomes]
\label{thm:resonant-signals-outcomes}
If $s_1 \res s_2$, then for each position~$i$, the interpretations
$r_i \compose s_1$ and $r_i \compose s_2$ resonate.
\end{theorem}

\begin{proof}[Proof sketch]
By induction on the repertoire, applying resonance-compose-left at each
step.  See \texttt{resonant\_signals\_resonant\_outcomes}.
\end{proof}

\subsection*{Discussion: Resonance as the Foundation of Comparative Anthropology}

Theorem~\ref{thm:resonant-signals-outcomes} has a profound methodological
implication: it \emph{justifies cross-cultural comparison}.  The
persistent anxiety of comparative anthropology---articulated most sharply
in Needham's \emph{Structure and Sentiment}~\cite{needham1962} and
renewed in the ``ontological turn'' of Viveiros de
Castro~\cite{viveirosdecastro2004}---is that comparison across cultures
is illegitimate because the ``same'' institution may mean entirely
different things in different societies.  Marriage among the Kachin is
not marriage among the English; sacrifice among the Nuer is not sacrifice
among the ancient Greeks.

The resonance theorem dissolves this anxiety by distinguishing between
\emph{identity} and \emph{resonance}.  Cross-cultural comparison does
not require that two cultural signals be identical; it requires only that
they \emph{resonate}---that they share sufficient compositional structure
to produce structurally related outcomes.  If the Roman Saturnalia and
modern Christmas are resonant signals ($s_{\text{Sat}} \res
s_{\text{Xmas}}$), then the theorem guarantees that for \emph{any}
receiver~$r_i$, the interpretations $r_i \compose s_{\text{Sat}}$ and
$r_i \compose s_{\text{Xmas}}$ will themselves resonate.  The comparison
is legitimate not because the festivals are ``the same'' but because
their compositional structures are sufficiently aligned to produce
structurally related interpretations.

\begin{example}[Christmas and Saturnalia]
\label{ex:christmas-saturnalia}
The structural resonance between the Roman Saturnalia and modern
Christmas has been noted since at least the early Church
Fathers.\footnote{See Miles~\cite{miles1912} for the classic historical
treatment and Nissenbaum~\cite{nissenbaum1996} for a social history of
how Christmas absorbed Saturnalian elements in the American context.}
Both involve: (i)~a period of festivity during the winter solstice;
(ii)~the temporary suspension or inversion of ordinary social
hierarchies (masters serving slaves; adults deferring to children);
(iii)~the exchange of gifts; (iv)~feasting and communal consumption.

These structural parallels are precisely the kind of shared compositional
features that ground resonance in IDT\@.  A Roman citizen composing
Saturnalia with their repertoire of ideas about social hierarchy,
reciprocity, and seasonal renewal produces an interpretation that
\emph{resonates with}---but is not identical to---the interpretation
produced by a modern American composing Christmas with their own
repertoire.  The theorem guarantees that this resonance is not
accidental but structural: it is a consequence of the compositional
alignment of the two signals.
\end{example}

% ── §6. Strategic Properties ─────────────────────────────────────────
\section{Strategic Properties}
\label{sec:ch2:strategic}

\begin{theorem}[Self-Signalling]
\label{thm:self-signalling}
If $\mathit{sender\_idea} \in \Rep$, then
$\mathit{sender\_idea} \compose s \in \operatorname{outcome}(G, s)$.
\end{theorem}

\begin{proof}[Proof sketch]
The sender's idea is mapped to its composition with~$s$ by the
interpretation function.  See \texttt{self\_signalling}.
\end{proof}

\begin{remark}
An insider---someone whose perspective is represented in the audience---
can always produce resonant interpretations by sending signals resonant
with their own position.
\end{remark}

\begin{theorem}[Composition Signal]
\label{thm:composition-signal}
\[
  \operatorname{outcome}(\Rep, s_1 \compose s_2)
  = \operatorname{outcome}(\operatorname{interpret}(\Rep, s_1), s_2).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By associativity of composition: for each~$r$,
$r \compose (s_1 \compose s_2) = (r \compose s_1) \compose s_2$.
See \texttt{composition\_signal}.
\end{proof}

\begin{remark}
A complex ritual (compound signal) has the same effect as first
``priming'' the audience with~$s_1$ and then delivering~$s_2$.  Ritual
structure exploits associativity.
\end{remark}

\begin{theorem}[Void Prepend]
\label{thm:void-prepend}
$\operatorname{outcome}(\Rep, \void \compose s)
 = \operatorname{outcome}(\Rep, s)$.
\end{theorem}

\begin{proof}[Proof sketch]
By void-right: $r \compose (\void \compose s) = r \compose s$ after
applying associativity and the void-right identity.
See \texttt{void\_prepend\_signal}.
\end{proof}

\begin{theorem}[Sequential Signal]
\label{thm:sequential-signal}
\[
  \operatorname{interpret}(\operatorname{interpret}(\Rep, s_1), s_2)
  = \operatorname{interpret}(\Rep, s_1 \compose s_2).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By functoriality of map and associativity.
See \texttt{sequential\_signal}.
\end{proof}

\begin{remark}
Multi-stage rituals (initiation sequences, liturgical calendars) produce
the same interpretive effect as a single complex ritual.  Ceremonial
structure is algebraically decomposable.
\end{remark}

\begin{example}[The Catholic Mass as Sequential Signal]
\label{ex:catholic-mass}
The Catholic Mass consists of a fixed sequence of signals: the
Introductory Rites (greeting, penitential act, \emph{Gloria}), the
Liturgy of the Word (readings, psalm, Gospel, homily), the Liturgy of
the Eucharist (offertory, consecration, communion), and the Concluding
Rites (blessing, dismissal).  By Theorem~\ref{thm:sequential-signal},
the interpretive effect of this four-part sequence is identical to
the effect of a single compound signal $s_{\text{intro}} \compose
s_{\text{word}} \compose s_{\text{eucharist}} \compose s_{\text{conclude}}$.

This equivalence has a practical consequence that liturgists have long
understood intuitively: the \emph{order} of the parts matters (because
composition is generally not commutative), but the \emph{division} into
parts is liturgically neutral (because sequential delivery equals
compound delivery).  A Mass celebrated with all four parts in sequence
produces exactly the same interpretive outcome as one in which the
entire rite were somehow delivered as a single act.\footnote{This
observation illuminates debates about liturgical reform.  The Second
Vatican Council's restructuring of the Mass (1962--1965) altered the
\emph{signals} but preserved the sequential structure.  By the
sequential signal theorem, the reform changed outcomes only insofar
as it changed the composed signal $s_1 \compose s_2 \compose s_3
\compose s_4$, not by virtue of reorganising the decomposition.}
\end{example}

\begin{remark}[Connections to Computer Science: Monadic Composition]
\label{rem:monadic}
The sequential signal theorem has a natural reading in terms of monadic
composition in functional programming.  The interpretation function
$\operatorname{interpret}(\Rep, -)$ acts as a \emph{Kleisli arrow} for
a list monad: it takes a signal and returns a list of ideas.  The
sequential signal theorem then states that Kleisli composition is
associative---that ``binding'' twice is the same as binding once with
the composed signal.  This is precisely the monad associativity law:
$(\mu \circ T\mu) = (\mu \circ \mu T)$.

For the computationally inclined reader, this correspondence is not
merely an analogy.  The Lean~4 formalisation exploits it directly:
the proof of \texttt{sequential\_signal} proceeds by rewriting with
\texttt{List.map\_map} (functoriality) and \texttt{compose\_assoc}
(associativity), which are exactly the two ingredients needed to
verify the monad laws.
\end{remark}

\begin{theorem}[Void Captures Signal]
\label{thm:void-captures}
If $\void \in G.\Rep$, then $s \in \operatorname{outcome}(G, s)$.
\end{theorem}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:negative-capability}.
See \texttt{void\_captures\_signal}.
\end{proof}

\begin{remark}
An ``open-minded'' receiver (one who retains $\void$ = pure receptivity)
always captures the sender's intended signal.
\end{remark}

\begin{theorem}[MaxDepth Outcome Bound]
\label{thm:max-outcome-depth}
\[
  \operatorname{maxInterpDepth}(\operatorname{outcome}(G, s))
  \leq \operatorname{maxInterpDepth}(G.\Rep) + \depth(s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:max-interp-depth}.
See \texttt{max\_outcome\_depth}.
\end{proof}

\begin{remark}
No matter how profound a signal, the deepest interpretation anyone can
produce is bounded by their existing depth plus the signal's depth.  The
receiver's capacity limits understanding, not the sender's intent---a
formalisation of ``you can't teach beyond the student's readiness.''
\end{remark}

\begin{remark}[Open Question: Tight Bounds]
\label{rem:tight-bounds}
Is the bound in Theorem~\ref{thm:max-outcome-depth} tight?  That is,
for every $d_R \geq 0$ and $d_S \geq 0$, does there exist a game~$G$
with $\operatorname{maxInterpDepth}(G.\Rep) = d_R$ and a signal~$s$ with
$\depth(s) = d_S$ such that
$\operatorname{maxInterpDepth}(\operatorname{outcome}(G,s)) = d_R + d_S$?
The answer is yes: take a repertoire containing a single idea of
depth~$d_R$ whose composition with any signal of depth~$d_S$ produces an
idea of depth exactly~$d_R + d_S$.  The ``leaf'' construction of
Chapter~1 provides such examples.  However, it remains open whether
tightness holds for \emph{natural} repertoires arising in ethnographic
practice, where the compositional structure of ideas may impose
additional constraints.  We conjecture that for ``generically chosen''
repertoires---those not specifically engineered to saturate the
bound---the actual maximum interpretation depth grows
sublinearly in $\depth(s)$.
\end{remark}

\begin{example}[A Socratic Dialogue as Bounded Signalling]
\label{ex:socratic}
In a Socratic dialogue, the teacher (Socrates) sends a sequence of
signals---questions---to a student.  Each question~$q_k$ has some
depth $\depth(q_k)$, and the student's repertoire evolves as
$\Rep_0 \to \Rep_1 \to \cdots$ via sequential interpretation.  By
Theorem~\ref{thm:max-outcome-depth}, the student's maximum
interpretation depth after question~$k$ satisfies
$\operatorname{maxInterpDepth}(\Rep_k) \leq
\operatorname{maxInterpDepth}(\Rep_{k-1}) + \depth(q_k)$.

Telescoping, the student's depth after $n$~questions is bounded by
$\operatorname{maxInterpDepth}(\Rep_0) + \sum_{k=1}^{n} \depth(q_k)$.
This formalises the Socratic insight that learning is \emph{incremental}:
the teacher cannot leap the student from depth~0 to depth~100 in a
single question, no matter how profound.  Each question contributes at
most its own depth to the student's understanding.  The ``art'' of
Socratic teaching consists in choosing questions whose depth is
calibrated to the student's current capacity---deep enough to extend
understanding, but not so deep as to be uninterpretable.\footnote{Plato's
\emph{Meno} dramatises exactly this calibration problem.  Socrates leads
the slave boy through a sequence of geometrical questions, each of which
extends the boy's understanding by a small increment.  The depth bound
theorem explains why this incremental strategy is not merely pedagogically
wise but \emph{mathematically necessary}.}
\end{example}

\begin{theorem}[Babbling Payoff]
\label{thm:babbling-payoff}
\[
  \Pay_S(\operatorname{outcome}(G, \void))
  = \Pay_S(G.\Rep).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By Theorem~\ref{thm:babbling-outcome}.
See \texttt{babbling\_payoff}.
\end{proof}

\begin{remark}
The babbling payoff equals the ``status quo'' payoff.  A signal is worth
sending only if it yields strictly more than silence.
\end{remark}

% ── Summary ──────────────────────────────────────────────────────────
\section*{Chapter Summary}

This chapter formalised the signalling problem within IDT\@.  The central
game-theoretic structure---a signalling game with repertoire-dependent
payoffs---was introduced, and the \emph{babbling equilibrium}
(Theorem~\ref{thm:babbling-outcome}) was established: silence always
leaves the receiver's mind unchanged.  We showed that outcomes have
invariant length (Theorem~\ref{thm:outcome-length}), bounded depth
(Theorem~\ref{thm:outcome-depth-bound}), and that resonant signals
produce resonant outcomes (Theorem~\ref{thm:resonant-signals-outcomes}).
The composition signal theorem (Theorem~\ref{thm:composition-signal})
reveals that compound signals decompose via associativity, providing the
algebraic foundation for analysing multi-stage rituals.

All eighteen theorems are verified in
\texttt{FormalAnthropology/IDT\_Signal\_Ch02.lean}.
