% ══════════════════════════════════════════════════════════════════════
% Chapter 5: Misinterpretation and Cultural Distance
% ══════════════════════════════════════════════════════════════════════

\chapter{Misinterpretation and Cultural Distance}
\label{ch:misinterp}

\epigraph{``Understanding is always more than merely re-creating someone
else's meaning.''}{Hans-Georg Gadamer, \textit{Truth and Method}}

% ── Introduction ──────────────────────────────────────────────────────

Every act of cross-cultural communication is haunted by the spectre of
misunderstanding.  The Sapir--Whorf hypothesis, in its strong form
(\emph{linguistic determinism}), holds that the structure of a language
determines the boundaries of its speakers' thought; in its weak form
(\emph{linguistic relativity}), it claims merely that language influences
perception and habitual thought.\footnote{The strong form, sometimes
called \emph{Whorfianism}, is largely discredited in contemporary
linguistics, but the weak form enjoys robust empirical support---see
Boroditsky~\cite{boroditsky2001} on how Mandarin speakers' vertical
metaphors for time measurably affect temporal reasoning.}  Both forms
share the premise that \emph{background shapes interpretation}, and it is
precisely this premise that Ideatic Depth Theory formalises through the
composition operator~$\compose$.

Walter Benjamin's luminous 1923 essay ``The Task of the Translator''
(\textit{Die Aufgabe des \"Ubersetzers}) argues that no translation can
reproduce the original's \emph{mode of meaning}
(\textit{Art des Meinens}): what is translatable is the literal content,
but what is lost is the way that content \emph{means} within its
original linguistic and cultural ecology.\footnote{Benjamin writes:
``Translation is so far removed from being the sterile equation of two
dead languages that of all literary forms it is the one charged with the
special mission of watching over the maturing process of the original
language and the birth pangs of its own.''  See
Benjamin~\cite{benjamin1923}, translated by Harry Zohn.}  In IDT terms,
Benjamin's insight is that the composition $r_{\mathrm{source}} \compose s$
and $r_{\mathrm{target}} \compose s$ may share propositional content
while diverging in ideatic depth---the ``mode of meaning'' is precisely
the depth structure that composition adds to the raw signal.

Eugene Nida, working on Bible translation in the mid-twentieth century,
drew a pragmatic distinction between \emph{formal equivalence} (preserving
the surface structure of the source text) and \emph{dynamic equivalence}
(preserving the reader's response).  IDT reveals these as two different
optimisation targets: formal equivalence minimises the distance between
$s_{\mathrm{source}}$ and $s_{\mathrm{target}}$, while dynamic
equivalence minimises the distance between
$r_{\mathrm{source}} \compose s_{\mathrm{source}}$ and
$r_{\mathrm{target}} \compose s_{\mathrm{target}}$---that is, between
the \emph{interpreted} signals rather than the raw ones.

Quine's \textit{Word and Object}~(1960) radicalises the problem further
with the thought experiment of \emph{radical translation}: a linguist
confronted with a wholly unknown language, with no bilingual informants,
must construct a translation manual from scratch.  Quine's famous
conclusion---the \emph{indeterminacy of translation}---states that
multiple, mutually incompatible translation manuals are equally
consistent with all possible behavioural evidence.\footnote{The
\emph{gavagai} example is canonical: when a native speaker points at a
rabbit and says ``gavagai,'' the linguist cannot determine whether the
word refers to the rabbit, to undetached rabbit-parts, or to a temporal
stage of rabbit-hood.  See Quine~\cite{quine1960}, Chapter~2.}  In IDT,
indeterminacy arises naturally: if $r_1 \compose s$ and $r_2 \compose s$
are both resonant with the observed behaviour, then the theory cannot
distinguish between backgrounds $r_1$ and $r_2$ on the basis of a
single signal~$s$ alone.

Vygotsky's developmental psychology and Cole's cross-cultural
experiments on cognition~\cite{cole1996} demonstrated that cognitive
processes are not universal but are \emph{culturally mediated}---that is,
the tools, symbols, and social practices of a culture shape how
individuals perceive, remember, and reason.  IDT provides the formal
scaffolding for this insight: the composition operator $r \compose s$ is
precisely the mechanism of cultural mediation, and the background~$r$
encodes the ``cultural toolkit'' through which the raw signal~$s$ is
processed.

This chapter formalises the intuition---shared by translators, diplomats,
anthropologists, and anyone who has ever struggled with a foreign
language---that \emph{something is always lost in translation}.  We give
that ``something'' a precise name (the interpretation gap), a formal
measure (resonance distance), and structural bounds (the depth
inequalities).

The sender intends the receiver to compose signal~$s$ with
background~$r^*$ (the ``shared understanding'').  But the receiver
actually composes with~$r_{\mathrm{actual}}$---their real, possibly
different, cultural background.  The gap between
$r^* \compose s$ and $r_{\mathrm{actual}} \compose s$ is the formal
measure of \emph{misinterpretation}.

This chapter develops the formal theory of interpretive divergence.
Connecting to the Sapir--Whorf hypothesis~\cite{vygotsky1978}, which
holds that linguistic background shapes interpretation, and to Gadamer's
hermeneutic circle~\cite{gadamer1960}, which insists that understanding
requires pre-understanding, we show that cultural distance between
backgrounds governs interpretive distance between outcomes.

\medskip
\noindent\textbf{Chapter overview.}  We define the interpretation gap
(\S\ref{sec:ch5:gap}), analyse void backgrounds
(\S\ref{sec:ch5:void-bg}), develop iterated re-interpretation
(\S\ref{sec:ch5:iteration}), study resonance under iteration
(\S\ref{sec:ch5:resonance-iter}), and formalise cultural distance
(\S\ref{sec:ch5:distance}).

\subsection*{Historical Context: From Babel to Formal Translation Theory}

The Tower of Babel narrative (Genesis~11:1--9) is the ur-text of
miscommunication theory: humanity once shared a single language and a
single interpretive background, and God's punishment was to shatter that
unity into mutually unintelligible fragments.  In IDT terms, the
pre-Babel state is one in which all agents share a common
background~$r^*$, so that $r_i = r^*$ for all~$i$ and every signal is
trivially well-interpreted.  The divine intervention replaces the single
background with a family $\{r_1, r_2, \ldots, r_k\}$ of pairwise
non-resonant backgrounds, guaranteeing systematic misinterpretation.

Friedrich Schleiermacher's 1813 lecture ``On the Different Methods of
Translating'' (\textit{\"Ueber die verschiedenen Methoden des
\"Ubersezens}) introduced a dichotomy that has structured translation
theory ever since: the translator must either bring the reader to the
author (\emph{foreignisation}) or bring the author to the reader
(\emph{domestication}).  Lawrence Venuti~\cite{venuti1995} revived this
distinction in the 1990s, arguing that the Anglo-American tradition
overwhelmingly favours domestication---a ``fluent'' translation that
effaces the foreignness of the source text.  In IDT, domestication
corresponds to replacing the source background~$r_{\mathrm{source}}$
with the target background~$r_{\mathrm{target}}$ before composing,
while foreignisation attempts to preserve $r_{\mathrm{source}}$ in the
composed output, accepting the resulting resonance gap as a feature
rather than a defect.

IDT provides a formal framework for measuring what is gained and lost
in cross-cultural interpretation.  Rather than relying on metaphors
(``faithful'' vs.\ ``free,'' ``spirit'' vs.\ ``letter''), we can state
precise theorems about how the structural properties of backgrounds---their
depth, their resonance relations, their compositional behaviour---govern
the structural properties of interpretive outcomes.  The remainder of
this chapter develops that theory.

% ── §1. The Interpretation Gap ───────────────────────────────────────
\section{The Interpretation Gap}
\label{sec:ch5:gap}

\begin{definition}[Intended Interpretation]
\label{def:intended-interp}
The sender's \emph{intended interpretation} is
$\operatorname{intendedInterp}(r_{\mathrm{intended}}, s)
 = r_{\mathrm{intended}} \compose s$.
\end{definition}

\begin{definition}[Actual Interpretation]
\label{def:actual-interp}
The receiver's \emph{actual interpretation} is
$\operatorname{actualInterp}(r_{\mathrm{actual}}, s)
 = r_{\mathrm{actual}} \compose s$.
\end{definition}

\begin{definition}[Well-Interpreted]
\label{def:well-interpreted}
An interpretation is \emph{well-interpreted} if the actual and intended
interpretations resonate:
\[
  \operatorname{wellInterpreted}(r_{\mathrm{int}}, r_{\mathrm{act}}, s)
  \;\iff\;
  r_{\mathrm{int}} \compose s \;\res\; r_{\mathrm{act}} \compose s.
\]
This is weaker than equality but stronger than ``no relation.''
\end{definition}

\begin{theorem}[Resonant Backgrounds Guarantee Well-Interpretation]
\label{thm:resonant-well-interpreted}
If $r_{\mathrm{int}} \res r_{\mathrm{act}}$, then the signal~$s$ is
well-interpreted.
\end{theorem}

\begin{proof}[Proof sketch]
By resonance-compose-right applied to the resonant backgrounds.
See \texttt{resonant\_backgrounds\_well\_interpreted}.
\end{proof}

\begin{remark}
Members of the same culture (resonant backgrounds) always produce
well-interpreted outcomes for any signal.  Gadamer's ``fusion of
horizons'' is guaranteed when horizons resonate~\cite{gadamer1960}.
\end{remark}

\begin{theorem}[Self-Interpretation Always Succeeds]
\label{thm:self-interpretation}
$\operatorname{wellInterpreted}(r, r, s)$ for all $r, s$.
\end{theorem}

\begin{proof}[Proof sketch]
By reflexivity of resonance on the backgrounds.
See \texttt{self\_interpretation}.
\end{proof}

\begin{remark}
You always understand yourself.  Soliloquy is the degenerate case of
communication.
\end{remark}

\begin{theorem}[Well-Interpretation is Symmetric]
\label{thm:well-interp-symm}
If $\operatorname{wellInterpreted}(r_1, r_2, s)$, then
$\operatorname{wellInterpreted}(r_2, r_1, s)$.
\end{theorem}

\begin{proof}[Proof sketch]
By symmetry of resonance.  See \texttt{well\_interpretation\_symm}.
\end{proof}

\begin{remark}
Mutual intelligibility is symmetric.  If I can understand your ritual,
you can understand mine.  Cultural exchange is inherently
bidirectional---a formalisation of the anthropological observation that
first contact is always reciprocal.
\end{remark}

\begin{example}[The ``Thumbs Up'' Gesture Across Cultures]
In the United States and much of Western Europe, a thumbs-up gesture is
a benign signal of approval or encouragement.  In parts of the Middle
East, West Africa, and South America, however, the same physical gesture
is considered obscene or deeply offensive.  Let $s$ denote the physical
signal (the raised thumb).  An American sender composes $s$ with the
background $r_{\mathrm{US}}$, which maps the gesture to the ideatic
content \emph{approval}.  A receiver in, say, Iraq composes $s$ with the
background $r_{\mathrm{IQ}}$, which maps the gesture to \emph{insult}.
Since $r_{\mathrm{US}} \compose s$ and $r_{\mathrm{IQ}} \compose s$
occupy entirely different regions of ideatic space, we have
$r_{\mathrm{US}} \compose s \not\res r_{\mathrm{IQ}} \compose s$---the
signal is \emph{not} well-interpreted, despite the sender's benign
intention.  The interpretation gap is not a failure of intelligence or
goodwill on either side; it is a structural consequence of non-resonant
backgrounds.
\end{example}

\begin{example}[Captain Cook's Arrival in Hawaii]
Marshall Sahlins's celebrated account of Captain James Cook's arrival in
Hawaii in January~1779 provides a dramatic illustration of the
interpretation gap at civilisational scale.\footnote{See
Sahlins~\cite{sahlins1985}, \textit{Islands of History}, Chapter~1.
Obeyesekere~\cite{obeyesekere1992} contests Sahlins's reading; the
debate itself is a meta-level instance of the interpretation gap.}  The
Hawaiians interpreted Cook's arrival through the schema of the god
Lono's return during the Makahiki festival: Cook's ships appeared from
the correct direction, at the correct time of year, and followed the
correct ritual circuit of the island.  Cook's intended signal---trade and
geographical exploration---was composed with a radically different
repertoire, $r_{\mathrm{Hawaiian}}$, that included the Lono myth as a
core component.  The resulting interpretation,
$r_{\mathrm{Hawaiian}} \compose s_{\mathrm{Cook}}$, was a divine
visitation, not a commercial expedition.  When Cook violated the ritual
script by returning unexpectedly after his departure (breaking the Lono
cycle), the interpretive framework collapsed, with fatal consequences.
\end{example}

\subsection*{Discussion: Symmetry and the Ethics of Cross-Cultural Understanding}

The symmetry theorem (Theorem~\ref{thm:well-interp-symm}) carries
ethical weight that extends beyond the purely formal.  If mutual
intelligibility is symmetric---if the fact that culture~$A$ can
understand culture~$B$'s signals implies that culture~$B$ can understand
culture~$A$'s signals---then the \emph{burden} of cross-cultural
understanding is necessarily shared.  Neither party has a structural
advantage or disadvantage in the interpretive exchange.

This result challenges colonial models of cultural encounter in which the
``primitive'' must learn to understand the ``civilised'' but not vice
versa.  The asymmetry posited by colonial discourse---that European
languages and categories are transparent while indigenous ones are
opaque---is formally incoherent within IDT\@.  If the missionary can
learn enough of the indigenous language to translate the Bible (i.e.,
$\operatorname{wellInterpreted}(r_{\mathrm{missionary}},
r_{\mathrm{indigenous}}, s_{\mathrm{Bible}})$), then the symmetry
theorem guarantees that the indigenous community can, in principle,
produce an equally valid interpretation of the missionary's cultural
artefacts.  The historical failure of colonial powers to recognise this
symmetry is not a consequence of structural asymmetry but of political
asymmetry---the refusal to grant interpretive authority to the colonised.

% ── §2. The Void Background ─────────────────────────────────────────
\section{The Void Background}
\label{sec:ch5:void-bg}

\begin{theorem}[Void Intended Transparency]
\label{thm:void-intended}
$\operatorname{intendedInterp}(\void, s) = s$.
\end{theorem}

\begin{proof}[Proof sketch]
By $\void \compose s = s$.  See \texttt{void\_intended\_transparency}.
\end{proof}

\begin{remark}
A sender who assumes zero shared context intends the signal to be
received ``as is.''  Any deviation from the raw signal is the receiver's
cultural contribution.  This is the formal model of the ``objective
journalist'' fantasy.
\end{remark}

\begin{theorem}[Void Actual Transparency]
\label{thm:void-actual}
$\operatorname{actualInterp}(\void, s) = s$.
\end{theorem}

\begin{proof}[Proof sketch]
By $\void \compose s = s$.  See \texttt{void\_actual\_transparency}.
\end{proof}

\begin{theorem}[Void-Void Perfect Interpretation]
\label{thm:void-void-perfect}
$\operatorname{wellInterpreted}(\void, \void, s)$ for all $s$.
\end{theorem}

\begin{proof}[Proof sketch]
Both interpret $s$ as $s$ itself, so the interpretations are identical
(and hence resonant).
See \texttt{void\_void\_perfect}.
\end{proof}

\begin{remark}
In the (impossible) limit of zero culture, there is zero
misinterpretation.  Misunderstanding is entirely a product of cultural
difference.
\end{remark}

\begin{theorem}[The Hermeneutic Circle]
\label{thm:hermeneutic-circle}
For any $r, s$:
$\operatorname{actualInterp}(r, s) \res \operatorname{actualInterp}(r, s)$.
\end{theorem}

\begin{proof}[Proof sketch]
By reflexivity of resonance.  See \texttt{hermeneutic\_circle}.
\end{proof}

\begin{remark}
Understanding always loops back to self-resonance.  You cannot escape
your own interpretive framework---even when you try to ``step outside,''
the result still resonates with your starting point.  This is the formal
content of Gadamer's hermeneutic circle~\cite{gadamer1960}.
\end{remark}

\begin{remark}[Connection to Psychoanalysis]
The hermeneutic circle (Theorem~\ref{thm:hermeneutic-circle}) has a
striking psychoanalytic analogue.  Freud's concept of
\emph{transference} (\textit{\"Ubertragung}) describes the process by
which the patient unconsciously interprets the analyst through schemas
derived from earlier relationships---typically parental ones.  The
patient's background $r_{\mathrm{patient}}$ includes these relational
schemas, and the ``signal'' $s_{\mathrm{analyst}}$ (the analyst's
words, silences, posture) is composed with them to produce an
interpretation that may have little to do with the analyst's actual
intentions.  Symmetrically, the analyst interprets the patient through
the theoretical background $r_{\mathrm{analytic}}$---Oedipal dynamics,
defence mechanisms, object relations---producing an interpretation that
is filtered through psychoanalytic training.  Both interpretations are
self-resonant (each is consistent with its own background), but they
may not resonate with each other.  The therapeutic process consists,
in IDT terms, of iteratively adjusting both backgrounds until mutual
resonance is achieved---a ``fusion of horizons'' in Gadamer's sense,
or a ``working alliance'' in the clinical vocabulary.
\end{remark}

\begin{example}[The Necker Cube and Perceptual Interpretation]
The Necker cube illusion demonstrates the hermeneutic circle at the
perceptual level.  The visual signal~$s$ is a two-dimensional line
drawing of a wireframe cube.  The human visual system composes $s$ with
a spatial schema~$r$ to produce a three-dimensional percept.  But there
are (at least) two non-resonant spatial schemas: $r_{\mathrm{front}}$,
which interprets the lower-left face as the front of the cube, and
$r_{\mathrm{back}}$, which interprets the same face as the back.  Each
interpretation is self-resonant---once you ``see'' the cube one way, all
the edges and vertices are consistent with that interpretation---but the
two interpretations do not resonate with each other.  The spontaneous
perceptual ``flip'' between the two readings is a switch between
self-resonant interpretive loops, not a correction of error.
\end{example}

\subsection*{Discussion: The Impossibility of the View from Nowhere}

Thomas Nagel's \textit{The View from Nowhere}~(1986) argues that
objectivity requires transcending one's particular perspective to
achieve a ``view from nowhere''---a standpoint that is not the standpoint
of any particular observer.\footnote{Nagel~\cite{nagel1986}
distinguishes between the ``subjective'' and ``objective'' poles of
understanding, arguing that complete objectivity is an ideal limit rather
than an achievable state.  IDT gives this intuition a formal backbone.}
The hermeneutic circle theorem provides a formal counterpart to Nagel's
scepticism about the attainability of such a standpoint.  Since every
interpretation is self-resonant---since
$\operatorname{actualInterp}(r, s) \res \operatorname{actualInterp}(r, s)$
for \emph{any} background~$r$---there is no composition that escapes the
gravitational pull of its own background.  The ``view from nowhere''
would require a background $r_{\varnothing}$ such that
$r_{\varnothing} \compose s = s$ for all~$s$; by
Theorem~\ref{thm:void-intended}, this is precisely the void
background~$\void$.  But the void background is not a ``view from
nowhere''---it is the view from \emph{no culture}, which, as we showed,
is the (impossible) limit of zero cultural content.  There is no
Archimedean point outside culture from which to evaluate interpretations
objectively; the best one can hope for is to make one's background
explicit and to seek resonance with other backgrounds through dialogue.

% ── §3. Iterated Re-Interpretation ───────────────────────────────────
\section{Iterated Re-Interpretation}
\label{sec:ch5:iteration}

\begin{definition}[Iterated Interpretation]
\label{def:iterate-interp}
The $n$-fold re-interpretation of signal~$s$ through background~$r$ is:
\[
  \operatorname{iterateInterp}(r, s, n) =
  \begin{cases}
    s & \text{if } n = 0, \\
    r \compose \operatorname{iterateInterp}(r, s, n-1) & \text{if } n > 0.
  \end{cases}
\]
This models rumination, echo chambers, and iterative cultural processing.
\end{definition}

\begin{lstlisting}[caption={Iterated interpretation in Lean~4},label=lst:iterate]
def iterateInterp (r s : I) : Nat -> I
  | 0 => s
  | n + 1 => IdeaticSpace.compose r (iterateInterp r s n)
\end{lstlisting}

\begin{theorem}[Zero Iteration = Raw Signal]
\label{thm:iterate-zero}
$\operatorname{iterateInterp}(r, s, 0) = s$.
\end{theorem}

\begin{proof}[Proof sketch]
By definition.  See \texttt{iterate\_zero}.
\end{proof}

\begin{theorem}[Single Iteration = Basic Interpretation]
\label{thm:iterate-one}
$\operatorname{iterateInterp}(r, s, 1) = r \compose s$.
\end{theorem}

\begin{proof}[Proof sketch]
By definition.  See \texttt{iterate\_one}.
\end{proof}

\begin{theorem}[Void Iteration Identity]
\label{thm:void-iterate}
$\operatorname{iterateInterp}(\void, s, n) = s$ for all $n$.
\end{theorem}

\begin{proof}[Proof sketch]
By induction on $n$, using $\void \compose x = x$ at the inductive step.
See \texttt{void\_iterate}.
\end{proof}

\begin{remark}
An ``empty mind'' processing a signal repeatedly never changes it.
Without cultural content, there is no interpretive drift.  Echo chambers
require content to echo.
\end{remark}

\begin{example}[Rumination and Depression]
Clinical psychology recognises that depressive rumination---the repeated,
involuntary processing of negative events through depressive
schemas---is a key maintaining mechanism of major depressive disorder.\footnote{
Nolen-Hoeksema~\cite{nolen-hoeksema1991} defines rumination as
``repetitively and passively focusing on symptoms of distress and on the
possible causes and consequences of these symptoms.''  The
response-styles theory predicts that rumination prolongs and intensifies
depressive episodes.}  In IDT terms, rumination is iterated
interpretation: $\operatorname{iterateInterp}(r_{\mathrm{depressive}},
s_{\mathrm{event}}, n)$, where $r_{\mathrm{depressive}}$ encodes the
cognitive distortions characteristic of depression (all-or-nothing
thinking, catastrophising, personalisation) and $s_{\mathrm{event}}$ is
the precipitating life event.  The depth bound theorem
(Theorem~\ref{thm:iterate-depth}) shows that this process cannot
generate unbounded complexity: after $n$ iterations, the depth of the
ruminated signal is at most
$n \cdot \depth(r_{\mathrm{depressive}}) + \depth(s_{\mathrm{event}})$.
The signal is transformed---often beyond recognition---but it remains
structurally bounded.  Cognitive-behavioural therapy intervenes not by
halting the iteration (which may be involuntary) but by replacing
$r_{\mathrm{depressive}}$ with a less distorting background
$r_{\mathrm{balanced}}$, thereby changing the trajectory of the orbit.
\end{example}

\begin{remark}[Connection to Dynamical Systems]
Iterated interpretation defines a discrete dynamical system on the
ideatic space~$\IS$.  Given a fixed background~$r$, the map
$T_r \colon \IS \to \IS$ defined by $T_r(x) = r \compose x$ generates
the orbit $\{s,\; r \compose s,\; r \compose (r \compose s),\; \ldots\}$
---a trajectory through ideatic space parameterised by the iteration
count~$n$.  The depth bound (Theorem~\ref{thm:iterate-depth}) constrains
this trajectory to a linearly growing region: the orbit cannot escape to
``ideatic infinity'' in finite time.  Whether the orbit converges to a
fixed point, enters a limit cycle, or wanders ergodically through a
bounded region remains an open question---one that connects IDT to the
rich literature on topological dynamics and symbolic dynamics.
\end{remark}

\begin{theorem}[Iterated Interpretation Depth Bound]
\label{thm:iterate-depth}
\[
  \depth(\operatorname{iterateInterp}(r, s, n))
  \leq n \cdot \depth(r) + \depth(s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By induction on $n$.  At step $n+1$, the depth-compose bound gives
$\depth(r \compose x) \leq \depth(r) + \depth(x)$; applying the
inductive hypothesis yields the result.
See \texttt{iterate\_depth\_bound}.
\end{proof}

\begin{remark}
Echo chambers have linearly bounded complexity.  Even infinite
rumination through a fixed cultural lens can only add linear depth.
Obsessive rumination is unproductive: each pass adds at most $\depth(r)$
more complexity.
\end{remark}

\begin{theorem}[Void Signal Single Iteration]
\label{thm:void-signal-iterate}
$\operatorname{iterateInterp}(r, \void, 1) = r$.
\end{theorem}

\begin{proof}[Proof sketch]
$r \compose \void = r$ by void-right.
See \texttt{void\_signal\_iterate\_one}.
\end{proof}

\begin{remark}
Processing ``nothing'' through a cultural lens yields the lens itself.
An empty news feed processed through conspiracy-theory schemas yields
the conspiracy schema unchanged.
\end{remark}

% ── §4. Resonance Under Iteration ────────────────────────────────────
\section{Resonance Under Iteration}
\label{sec:ch5:resonance-iter}

\begin{theorem}[Resonant Iteration Step]
\label{thm:resonant-step}
If $r_1 \res r_2$ and $x_1 \res x_2$, then
$r_1 \compose x_1 \;\res\; r_2 \compose x_2$.
\end{theorem}

\begin{proof}[Proof sketch]
By the full resonance-compose lemma.
See \texttt{resonant\_iteration\_step}.
\end{proof}

\begin{theorem}[Full Resonant Iteration]
\label{thm:full-resonant-iteration}
If $r_1 \res r_2$, then for all $n$:
\[
  \operatorname{iterateInterp}(r_1, s, n)
  \;\res\;
  \operatorname{iterateInterp}(r_2, s, n).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By induction on $n$.  Base case: $s \res s$ by reflexivity.  Inductive
step: apply Theorem~\ref{thm:resonant-step} with the inductive
hypothesis.
See \texttt{resonant\_iteration}.
\end{proof}

\begin{remark}
Cultural similarity guarantees interpretive similarity at every stage
of processing.  This is why ``deep'' cultural understanding between
allied groups is possible---resonance propagates through all layers of
interpretation.
\end{remark}

\begin{theorem}[Self-Iteration Self-Resonance]
\label{thm:self-iter-resonance}
$\operatorname{iterateInterp}(r, s, n) \res
 \operatorname{iterateInterp}(r, s, n)$ for all $r, s, n$.
\end{theorem}

\begin{proof}[Proof sketch]
By reflexivity.  See \texttt{self\_iteration\_resonance}.
\end{proof}

\begin{remark}
The echo chamber's output is always ``consistent'' in the sense of
self-resonance---it just may not resonate with \emph{anything else}.
This is the formal trap of intellectual isolation.
\end{remark}

\begin{example}[Protestant and Catholic Interpretations of the Eucharist]
Protestant and Catholic Christians process the Eucharist---the central
ritual of Christian worship---through different theological backgrounds.
The Catholic background $r_{\mathrm{Catholic}}$ includes the doctrine of
\emph{transubstantiation}: the bread and wine literally become the body
and blood of Christ, even though their sensory properties
(\emph{accidents}) remain unchanged.  The mainline Protestant background
$r_{\mathrm{Protestant}}$ includes a \emph{symbolic memorial} reading:
the bread and wine \emph{represent} Christ's sacrifice but do not
undergo ontological transformation.\footnote{The spectrum is, of course,
more complex: Luther's \emph{consubstantiation}, Zwingli's pure
memorialism, and Calvin's ``spiritual real presence'' occupy intermediate
positions.  We simplify here for illustrative purposes.}  Let $s$ denote
the ritual signal (the words of institution, the breaking of bread, the
pouring of wine).  The two interpretations,
$r_{\mathrm{Catholic}} \compose s$ and
$r_{\mathrm{Protestant}} \compose s$, differ substantially in their
ontological commitments.  Nevertheless, since
$r_{\mathrm{Catholic}} \res r_{\mathrm{Protestant}}$---both share the
core Christian schemas of incarnation, sacrifice, and redemption---Theorem~\ref{thm:full-resonant-iteration} guarantees that their
iterated interpretations remain resonant at every stage of theological
elaboration.  The disagreement is real but structurally bounded: both
traditions can recognise the other's Eucharistic practice as a genuine,
if mistaken, interpretation of the same signal.
\end{example}

\subsection*{Discussion: Echo Chambers as Iterated Self-Interpretation}

The connection between iterated re-interpretation and echo chambers
(developed more fully in Chapter~\ref{ch:echo}) deserves preliminary
comment here.  When a community processes information exclusively
through its own interpretive lens---when Fox News viewers watch only Fox
News, or when members of a conspiracy-theory forum read only posts from
other members---the community performs iterated self-interpretation:
$\operatorname{iterateInterp}(r_{\mathrm{community}}, s, n)$ for
increasing~$n$.

The self-resonance theorem (Theorem~\ref{thm:self-iter-resonance}) shows
that the output is always internally consistent: at every iteration, the
signal resonates with itself.  But \emph{consistency is not the same as
accuracy}.  A conspiracy theory is self-resonant---every new piece of
evidence is absorbed into the conspiratorial framework without
contradiction---but it may not resonate with any signal originating
outside the community's background.  The formal distinction between
self-resonance (guaranteed) and cross-resonance (contingent) is the key
to understanding why echo chambers feel so convincing from the inside
and so baffling from the outside.

% ── §5. Cultural Distance ────────────────────────────────────────────
\section{Cultural Distance and Interpretation}
\label{sec:ch5:distance}

\begin{theorem}[Composition Preserves Well-Interpretation]
\label{thm:compound-well-interpreted}
If $s_1$ and $s_2$ are both well-interpreted under the same background
pair $(r_1, r_2)$, then
\[
  (r_1 \compose s_1) \compose (r_1 \compose s_2)
  \;\res\;
  (r_2 \compose s_1) \compose (r_2 \compose s_2).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By the full resonance-compose lemma applied to the two well-interpretation
hypotheses.
See \texttt{compound\_well\_interpreted}.
\end{proof}

\begin{remark}
If two cultures understand each other's simple signals, they understand
compound signals too.  Mutual intelligibility at the atomic level
guarantees mutual intelligibility at the molecular level.  This is why
pidgins can bootstrap into creoles.
\end{remark}

\begin{theorem}[Interpretation Depth Gap]
\label{thm:interp-depth-gap}
\[
  \depth(\operatorname{actualInterp}(r_{\mathrm{actual}}, s))
  \leq \depth(r_{\mathrm{actual}}) + \depth(s).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By depth-compose-le.  See \texttt{interpretation\_depth\_gap}.
\end{proof}

\begin{remark}
Misinterpretation does not create unbounded complexity.  Even the most
``wrong'' interpretation is bounded by the receiver's actual depth plus
the signal's depth.
\end{remark}

\begin{theorem}[Iteration Composition / Ritual Accretion]
\label{thm:iteration-composition}
\[
  \operatorname{iterateInterp}(r, s, n + m)
  = \operatorname{iterateInterp}(r,\;
    \operatorname{iterateInterp}(r, s, m),\; n).
\]
\end{theorem}

\begin{proof}[Proof sketch]
By induction on $n$.  See \texttt{iteration\_composition}.
\end{proof}

\begin{remark}
A lifetime of cultural processing ($n + m$ iterations) decomposes into
childhood processing ($m$~iterations) followed by adult processing
($n$~iterations).  Cultural development is \emph{sequentially
decomposable}: we can study ``what did early formation contribute?'' and
``what did later experience add?'' as separable phases.  This is the
formal basis of Bourdieu's distinction between primary and secondary
habitus formation~\cite{bourdieu1984}.
\end{remark}

\begin{remark}[Connection to Developmental Psychology]
Piaget's stage theory of cognitive development can be read through the
lens of iteration composition.  The sensorimotor stage (ages~0--2)
establishes a foundational background $r_{\mathrm{sensorimotor}}$
through which all subsequent signals are processed.  The preoperational
stage (ages~2--7) iterates on top of this foundation, producing
$\operatorname{iterateInterp}(r_{\mathrm{preop}},
\operatorname{iterateInterp}(r_{\mathrm{sensorimotor}}, s, m_1), m_2)$.
The decomposability theorem (Theorem~\ref{thm:iteration-composition})
formalises Piaget's insight that cognitive development is sequential but
analytically decomposable: each stage builds on the previous one, and
the total interpretation after all stages equals the sequential
composition of the individual stages.  This does \emph{not} mean that
the stages are independent---the output of the sensorimotor stage
becomes the input to the preoperational stage---but it does mean that
the contribution of each stage can be isolated and studied.
\end{remark}

\begin{example}[Language Acquisition as Iterated Interpretation]
A child acquiring language iteratively interprets speech sounds through
developing phonological schemas.  In the first months of life, the
infant's phonological background $r_{\mathrm{phon}}^{(0)}$ is
language-universal: all human languages' phonemic contrasts are
discriminable.\footnote{Kuhl~\cite{kuhl2004} demonstrates that infants
under six months can discriminate phonemic contrasts from any human
language, but this ability narrows to the native language's contrasts by
twelve months---a process called ``perceptual narrowing'' or ``neural
commitment.''}  By twelve months, iterated exposure to the ambient
language has specialised $r_{\mathrm{phon}}^{(0)}$ into a
language-specific background $r_{\mathrm{phon}}^{(n)}$ that enhances
native contrasts and suppresses non-native ones.  The depth bound on
iterated interpretation (Theorem~\ref{thm:iterate-depth}) explains why
language acquisition follows a predictable trajectory: each iteration
adds at most $\depth(r_{\mathrm{phon}})$ to the child's linguistic
complexity, so the growth of phonological competence is at most linear in
the number of exposures.
\end{example}

\begin{remark}[Open Questions]
Several questions remain open within the framework developed in this
chapter.  First: does iterated interpretation always converge to a fixed
point?  That is, does there exist, for every $r$ and~$s$, an
$N \in \mathbb{N}$ such that
$\operatorname{iterateInterp}(r, s, N) = \operatorname{iterateInterp}(r, s, N+1)$?
Second: under what conditions does the orbit
$\{s,\; r \compose s,\; r \compose (r \compose s),\; \ldots\}$
stabilise, and when does it cycle?  These questions connect to the
fixed-point theory developed in Chapter~\ref{ch:echo} and to the
broader question of whether cultural evolution has attractors.  Third:
can the depth bound (Theorem~\ref{thm:iterate-depth}) be tightened?
If the composition operator satisfies additional structural
properties---subadditivity, contractivity---then the linear bound may be
improvable to a logarithmic or even constant bound, with significant
implications for the theory of echo chambers and cultural convergence.
\end{remark}

% ── Summary ──────────────────────────────────────────────────────────
\section*{Chapter Summary}

This chapter developed the formal theory of misinterpretation within
IDT\@.  The \emph{interpretation gap} between intended and actual
interpretation is governed by the resonance between sender and receiver
backgrounds (Theorem~\ref{thm:resonant-well-interpreted}).
Well-interpretation is symmetric
(Theorem~\ref{thm:well-interp-symm}) and guaranteed for identical
backgrounds (Theorem~\ref{thm:self-interpretation}).  \emph{Iterated
re-interpretation} has linearly bounded depth
(Theorem~\ref{thm:iterate-depth}), and \emph{resonant iteration}
(Theorem~\ref{thm:full-resonant-iteration}) shows that culturally
similar backgrounds produce resonant interpretations at every stage.
The \emph{iteration composition} theorem
(Theorem~\ref{thm:iteration-composition}) provides the formal basis for
decomposing cultural development into sequential phases.

This concludes Part~I\@.  We have established the interpretive
framework: repertoires, signalling games, costly signals, resonance
activation, and misinterpretation.  Part~II develops the equilibrium
theory.

All eighteen theorems are verified in
\texttt{FormalAnthropology/IDT\_Signal\_Ch05.lean}.
